### 2024-05-05 doc/nine-species/nine-species.tex
Add starter files

\documentclass{article}
\usepackage{graphicx} % Allow insertion of graphics.
\usepackage{authblk} % Better layout of affiliations.
\usepackage[margin=1in]{geometry} % Set margins.
\usepackage{xcolor} % Allow colored text.
\usepackage{url} % Allow URLs.
\usepackage[sort&compress]{natbib} % Better bibliography formatting.

\newcommand{\fixme}[1]{{\color{red}{\bf FIXME: #1}\color{black}}}

\title{A multi-species benchmark for training and validating large scale mass spectrometry proteomics machine learning models}

\author[1,2]{William Stafford Noble}

\affil[1]{Department of Genome Sciences, University of Washington}
\affil[2]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington}

\date{}

\begin{document}

\maketitle

\fixme{Instructions at
  \url{https://www.nature.com/sdata/publish/submission-guidelines}.}

\begin{abstract}
  \fixme{Insert abstract here.}
\end{abstract}

\section*{Background \& Summary}

Similar to many other fields of scientific inquiry, the analysis of proteomics tandem mass spectrometry data is being revolutionized by the development of deep learning analysis methods.   For example, deep learning has been used to perform large-scale embedding and clustering of peptide mass spectra \cite{bittremieux2022learned}, predict fragment ion intensities \cite{zhou2017pdeep, gessulat2019prosit, tiwary2019high}, predict peptide liquid chromatography retention times \fixme{\cite{XXX}}, identify peptide features in precursor (MS1) spectra \cite{zohora2019deepiso, zohora2021deep}, and perform \textit{de novo} peptide sequencing \cite{tran2017denovo, qiao2021computationally, yilmaz2022denovo, liu2023accurate, mao2023mitigating, klaproth2024deep, lee2024bidirectional}.

Training and validation data for these models are typically obtained from two types of sources: public repositories of proteomics mass spectrometry data, such as PRIDE \cite{martens2005pride} and MassIVE \fixme{\cite{XXX}}, or the ProteomeTools repository of data derived from synthesized peptide seuqences \cite{zolg2017building}.  In the former case, further processing of data from the public repositories may be necessary to create a consistently annotated collection of spectra, as was done for the MassIVE-KB dataset of annotated human spectra \cite{wang2018assembling}.

Here we describe a large collections of proteomics tandem mass spectrometry data that have been collated and formatted for use in training and validating large-scale machine learning models.  All of the spectra in these datasets are MS2 spectra, meaning that each spectrum nominally represents peaks produced from the fragmention of a single peptide species.  Thus, the primary data object is a peptide-spectrum match (PSM), consisting of a single MS2 spectrum and its associated peptide sequence and charge state.  The second dataset, the nine-species benchmark, is derived from the PRIDE repository and consists of 2.8 million PSMs and 13 million unannotated spectra from nine different species.  A version of this benchmark was originally described in the DeepNovo paper \cite{trans2017denovo}.  The version reported here was re-created from the original PRIDE datasets using a different search engine, a wider range of post-translational modifications, and enforcing segregation of peptides between the different species in the benchmark.  This dataset has been used in the training and validation of the Casanovo \textit{de novo} peptide sequencing model \cite{yilmaz2022denovo}.

\section*{Methods}

\subsection*{Nine-species benchmark}

\fixme{I need to rewrite this paragraph to avoid self-plagiarism. ---Bill}
\begin{sloppypar}
We created a new version of the nine-species benchmark originally described by Tran {\em et al.} \cite{tran2017denovo}
To do so, we downloaded the RAW files from the same nine PRIDE projects (Supplementary Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4.
We also downloaded the corresponding nine Uniprot reference proteomes and constructed a Tide index for each one, using Crux version 4.1.
For one species (\textit{Vigna mungo}), no reference proteome is available, so we used the proteome of the closely related species \textit{Vigna radiata}.
We specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
Each index also contains a shuffled decoy peptide corresponding to each target peptide.
Each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Supplementary Table~\ref{tab:benchmark}).
We used XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
All search results were then analyzed jointly per species using the Crux implementation of Percolator, with default parameters.
For the benchmark, we retained all PSMs with Percolator q value $<$ 0.01.
We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
We then post-processed the PSMs to eliminate peptides that are shared between species.
Among the 229,984 unique peptides, we identified 3797 (1.7\%) that occur in more than one species.
For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
Note that when identifying shared peptides between species, we
considered all modified forms of a given peptide sequence to be the
same.  Hence, if a given peptide appears in more than one species,
then that peptide, including all its modified forms, is randomly
assigned to a single species and eliminated from the others.
The final benchmark dataset consists of 2.8 million PSMs drawn from 343 RAW files.
The revised nine-species benchmark is available on MassIVE at \url{https://doi.org/doi:10.25345/C52V2CK8J}.
\end{sloppypar}

\section*{Data Records}

\fixme{
\begin{itemize}
\item Annotated MGF files for the nine-species benchmark. \fixme{Do we provide them separately by run, or just record in the MGF records what run each one came from?}
\item Intermediate files for nine-species (genome FASTA, tide-search results, Percolator files).
\end{itemize}
}

\section*{Technical Validation}

\section*{Usage Notes (optional)}

\section*{Code Availability}

\fixme{I need to cobble together a couple of scripts to generate the pipeline, based on \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2024-02-13deepnovo9/runall}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2024-02-13deepnovo9/summarize}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2024-02-13deepnovo9/deepnovo.txt}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2023-01-19deepnovo7/runall}, and \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/bin/clean-benchmark.py}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/bin/annotate_mgf.py}. ---Bill}

\section*{References}

\paragraph{Author Contributions}

\paragraph{Competing Interests}

\bibliographystyle{unsrt}
\bibliography{refs} % See https://github.com/Noble-Lab/noble-lab-references

\section*{Supplementary information}

\appendix
\renewcommand{\theequation}{S\arabic{equation}}
\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thesection}{S\arabic{section}}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\setcounter{figure}{0}

\end{document}

% How to make a dense list.
%\usepackage{enumitem}
%\begin{enumerate}[noitemsep,nolistsep]

### 2024-05-12 doc/nine-species/nine-species.tex
Various updates

@@
 
-Similar to many other fields of scientific inquiry, the analysis of proteomics tandem mass spectrometry data is being revolutionized by the development of deep learning analysis methods.
-For example, deep learning has been used to perform large-scale embedding and clustering of peptide mass spectra \cite{bittremieux2022learned}, predict fragment ion intensities \cite{zhou2017pdeep, gessulat2019prosit, tiwary2019high}, predict peptide liquid chromatography retention times \fixme{\cite{XXX}}, identify peptide features in precursor (MS1) spectra \cite{zohora2019deepiso, zohora2021deep}, and perform \textit{de novo} peptide sequencing \cite{tran2017denovo, qiao2021computationally, yilmaz2022denovo, liu2023accurate, mao2023mitigating, klaproth2024deep, lee2024bidirectional}.
+\textit{De novo} sequencing of proteomics tandem mass spectrometry data, in which observed fragmentation spectra are translated into corresponding peptide sequences, has been an open challenge for more than 40 years \cite{sakurai1984paas}.
+Recently, as in many other areas of science, considerable progress toward solving this challenge has been made using deep learning, in which multi-layer neural networks with millions of parameters are trained to generate peptide sequences from observed spectra.
+The first such deep learning method, DeepNovo \cite{tran2017denovo}, has been followed by at least 22 additional publications (reviewed in \cite{XXX}).
 
-Training and validation data for these models are typically obtained from two types of sources: public repositories of proteomics mass spectrometry data, such as PRIDE \cite{martens2005pride} and MassIVE \fixme{\cite{XXX}}, or the ProteomeTools repository of data derived from synthesized peptide seuqences \cite{zolg2017building}.
-In the former case, further processing of data from the public repositories may be necessary to create a consistently annotated collection of spectra, as was done for the MassIVE-KB dataset of annotated human spectra \cite{wang2018assembling}.
+Because deep learning models are large, they typically require a correspondingly large amount of data to train.
+\fixme{Mention how many parameters some important models have, and how much data they were trained on.}
 
-Here we describe a large collections of proteomics tandem mass spectrometry data that have been collated and formatted for use in training and validating large-scale machine learning models.
-All of the spectra in these datasets are MS2 spectra, meaning that each spectrum nominally represents peaks produced from the fragmention of a single peptide species.
-Thus, the primary data object is a peptide-spectrum match (PSM), consisting of a single MS2 spectrum and its associated peptide sequence and charge state.
-The second dataset, the nine-species benchmark, is derived from the PRIDE repository and consists of 2.8 million PSMs and 13 million unannotated spectra from nine different species.
-A version of this benchmark was originally described in the DeepNovo paper \cite{trans2017denovo}.
-The version reported here was re-created from the original PRIDE datasets using a different search engine, a wider range of post-translational modifications, and enforcing segregation of peptides between the different species in the benchmark.
-This dataset has been used in the training and validation of the Casanovo \textit{de novo} peptide sequencing model \cite{yilmaz2022denovo}.
+\fixme{Talk about complications.  Labeling strategy.  Train/test leakage.  Batch effects.  Quality of the spectra.}
 
@@
 
-\subsection*{Nine-species benchmark}
+\subsection*{Data sets}
 
-\fixme{I need to rewrite this paragraph to avoid self-plagiarism. ---Bill}
-\begin{sloppypar}
+\fixme{Describe each study}
+
 We created a new version of the nine-species benchmark originally described by Tran {\em et al.} \cite{tran2017denovo}
@@
 For one species (\textit{Vigna mungo}), no reference proteome is available, so we used the proteome of the closely related species \textit{Vigna radiata}.
+
+\subsection*{Assigning peptide labels to spectra}
+
 We specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
@@
 We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
+
+\subsection*{Avoiding train/test leakage}
+
 We then post-processed the PSMs to eliminate peptides that are shared between species.
@@
 The final benchmark dataset consists of 2.8 million PSMs drawn from 343 RAW files.
-The revised nine-species benchmark is available on MassIVE at \url{https://doi.org/doi:10.25345/C52V2CK8J}.
-\end{sloppypar}
+
+\subsection*{Balancing the benchmark}
+
+At this stage, the benchmark is quite imbalanced, in the sense that some species have a much larger number of associated PSMs.
+We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
+Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,541.
+Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
+To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
+This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,604 for \textit{H.\ sapiens} and 82,290 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
+Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
+The final, balanced benchmark dataset consists of \fixme{XXX} PSMs.
 
@@
 
+\fixme{False discovery rate curves}
+
 \section*{Usage Notes (optional)}
@@
 \section*{Code Availability}
-
-\fixme{I need to cobble together a couple of scripts to generate the pipeline, based on \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2024-02-13deepnovo9/runall}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2024-02-13deepnovo9/summarize}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2024-02-13deepnovo9/deepnovo.txt}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/results/bill/2023-01-19deepnovo7/runall}, and \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/bin/clean-benchmark.py}, \url{https://github.com/Noble-Lab/2021_melih_ms-chimera/blob/main/bin/annotate_mgf.py}.
----Bill}
 

### 2024-05-12 bin/summarize_benchmark.py

@@ -1,0 +1,101 @@
+#!/usr/bin/env python
+# AUTHOR: William Stafford Noble
+# CREATE DATE: 10 May 2024, Branford, CT
+import sys
+import argparse
+import pandas
+import os
+import glob
+
+DESCRIPTION = """Write text and HTML tables describing a given
+benchmark dataset."""
+
+def count_spectra(mgf_filenames):
+    "Count number of spectra in a set of MGF files."
+    return_value = 0
+
+    for mgf_filename in mgf_filenames:
+        with open(mgf_filename, "r") as mgf_file:
+            for line in mgf_file:
+                if line.rstrip() == "BEGIN IONS":
+                    return_value += 1
+
+    return return_value
+
+def count_lines(filename):
+    with open(filename, "r") as my_file:
+        return(len(my_file.readlines()))
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+    global DESCRIPTION
+
+    # Parse the command line.
+    parser = argparse.ArgumentParser(description=DESCRIPTION)
+    parser.add_argument('--driver_filename', type=str, required=True,
+                        help="Name of driver file.  Columns are PXD ID, "
+                        + "species name, precursor tolerance (ppm), fragment "
+                        + "bin size (m/z)")
+    parser.add_argument('--data_dir', type=str, required=True,
+                        help="Directory containing original data")
+    parser.add_argument('--benchmark_dir', type=str, required=True,
+                        help="Directory containing benchmark")
+    parser.add_argument('--root', type=str, required=True, 
+                        help="Output directory and filename root")
+    args = parser.parse_args()
+
+    driver = pandas.read_csv(args.driver_filename, sep="\t", header=None)
+    all_species = list(driver[1])
+
+    column_names = ["#raw", "#mgf", "#spectra", "#PSMs", "#peptides"]
+    output = {} # Key = column name, value = list of entries.
+    output["species"] = all_species
+    for stat in column_names:
+        output[stat] = []
+
+    # Gather up all the statistics
+    for species in all_species:
+        for stat in column_names:
+            if stat == "#raw":
+                output[stat].append(
+                    len(glob.glob(os.path.join(args.data_dir,
+                                               species, "*.mgf")))
+                )
+            elif stat == "#mgf":
+                output[stat].append(
+                    len(glob.glob(os.path.join(args.benchmark_dir,
+                                               species, "*.mgf")))
+                )
+            elif stat == "#spectra":
+                output[stat].append(
[…]

### 2024-05-13 bin/downsample_benchmark.py
Add downsampling script

@@ -1,0 +1,64 @@
+#!/usr/bin/env python
+# AUTHOR: William Stafford Noble
+# CREATE DATE: 10 May 2024, Branford, CT
+import sys
+import argparse
+import random
+import pyteomics.mgf
+import os
+import glob
+import shutil
+
+DESCRIPTION = """Select from a collection of sets of MGF files so that
+each set has approximately the same number of spectra. This is done by
+randomly permuting the order of MGFs within each set and then
+accepting files in order until at least a specified number of spectra
+is obtained."""
+
+def count_spectra(mgf_filename):
+    """Count the number of spectra in an MGF file."""
+    return(len(pyteomics.mgf.read(mgf_filename)))
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+    global DESCRIPTION
+
+    # Parse the command line.
+    parser = argparse.ArgumentParser(description=DESCRIPTION)
+    parser.add_argument('mgf_dirs', nargs='+', help="MGF directories")
+    parser.add_argument('--num_spectra', type=int, required=False,
+                        default=100000, help="Target number of spectra per set")
+    parser.add_argument('--seed', type=int, required=False, default=7718,
+                        help="Random number seed")
+    parser.add_argument('--root', type=str, required=True, 
+                        help="Output directory name")
+    args = parser.parse_args()
+
+    random.seed(args.seed)
+    os.makedirs(args.root, exist_ok=True)
+
+    for mgf_dir in args.mgf_dirs:
+        os.makedirs(os.path.join(args.root, os.path.basename(mgf_dir)),
+                    exist_ok=True)
+        mgf_filenames = glob.glob(os.path.join(mgf_dir, "*.mgf"))
+        random.shuffle(mgf_filenames)
+        total_spectra = 0
+        for mgf_filename in mgf_filenames:
+            new_filename = os.path.join(args.root,
+                                        os.path.basename(mgf_dir),
+                                        os.path.basename(mgf_filename))
+            shutil.copy(mgf_filename, new_filename)
+            num_spectra = count_spectra(mgf_filename)
+            print(f"Read {num_spectra} from {mgf_filename}.")
+            total_spectra += num_spectra
+            if (total_spectra > args.num_spectra):
+                print(f"Limit reached ({total_spectra} > {args.num_spectra})"
+                      + f" for {mgf_dir}.",
+                      file=sys.stderr)
+                break
+
+if __name__ == "__main__":
+    main()
+    sys.exit(0)

### 2024-05-15 doc/nine-species/nine-species.tex
Add descriptions of studies

@@
 
-\fixme{Describe each study}
+We created a new version of the nine-species benchmark originally described by Tran {\em et al.} \cite{tran2017denovo}
 
-We created a new version of the nine-species benchmark originally described by Tran {\em et al.} \cite{tran2017denovo}
+\begin{enumerate}
+\item Paiva \textit{et al.}\ investigated the protein expression response of the cowpea plant (\textit{Vigna unguiculata}) to infection by \textit{Cowpsea severe mosaic virus} (CSMV) by carrying out label-free proteomic analysis of cowpea leaves that were inoculated with CSMV compared to mock inoculation controls \cite{paiva2016label}.
+\item Nevo \textit{et al.}\ studied a rare autosomal recessive lysosomal storage disorder, cystinosis, by carrying out SILAC proteomic analysis of engineered mouse cell lines that harbor a known pathenogenic mutation of the causative gene, \textit{CTNS} \cite{nevo2017impact}.
+\item Cassidy \textit{et al.}\ evaluated two different analytical approaches for carrying out full proteome analysis while identifying short open reading frames: high/low pH reversed phase LC-MS bottom-up approach and a semi-top-down strategy involving separation of proteins a GelFree system followed by digestion and LC-MS analysis \cite{cassidy2016combination}.
+The experiments were carried out using the methane producing archaeon \textit{Methanosarcina mazei}.
+\item Reuss \textit{et al.}\ carried out proteomic analyses on a series of minimized strains of the model bacterium, \textit{Bacillus subtilis}, with genomes reduced by $\sim$36\% \cite{reuss2017large}.
+\item Petersen \textit{et al.}\ performed proteomic analysis of \textit{Candidatus endoloripes}, which are bacterial symbionts of the \textit{Lucinidae} family of marine bivalves \cite{petersen2016chemosynthetic}.
+\item Mata \textit{et al.}\ characterized the proteome of the tomato pericarp at its ripe red stage \cite{mata2017depth}.
+\item Seidel \textit{et al.}\ analyzed the global proteomic stress response in wildtype and two yeast knockout strains for the gene PBP1  \cite{seidel2017quantitative}.
+\item Hu \textit{et al.}\ studied honeybees that exhibit a suite of behaviors (\textit{Varroa} sensitive hygiene---VSH) associated with infection with the \textit{Varroa destructor} virus \cite{hu2016proteome}.
+Proteomic analysis was carried out on mushroom bodies and antennae of adult honeybees with and without VSH.
+\item Cypryck \textit{et al.} characterized extracellular vesicles released from human primary macrophages after infection with influenza A viruses \cite{cypryk2017proteomic}.
+\end{enumerate}
+All nine studies were performed using a Thermo Scientific Q Exactive mass spectrometer.
+
+% The PXD entry says the species is "Vigna mungo (Rice bean) (Black gram); NCBI TaxID: 3915;" but the paper talks about Vinga unguiculata.
+
 To do so, we downloaded the RAW files from the same nine PRIDE projects (Supplementary Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4.
@@
 
-\fixme{
+The dataset contains files resulting from various stages of the benchmark:
 \begin{itemize}
-\item Annotated MGF files for the nine-species benchmark. \fixme{Do we provide them separately by run, or just record in the MGF records what run each one came from?}
-\item Intermediate files for nine-species (genome FASTA, tide-search results, Percolator files).
+\item Spectrum files in MGF format, produced by msconvert.
+\item Reference proteome files in FASTA format, downloaded from Uniprot.
+\item Search results files for both targets and decoys, in tab-delimited format, produced by Tide.
+\item PSM-level Percolator results files for targets, in tab-delimited format.
+\item Annotated MGF files for all three versions of the benchmark (overlap, main, and balanced).
 \end{itemize}
-}
+Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
 

### 2024-07-11 doc/nine-species/nine-species.tex
Edits

@@
 
-Because deep learning models are large, they typically require a correspondingly large amount of data to train.
-\fixme{Mention how many parameters some important models have, and how much data they were trained on.}
+The standard method for evaluating these \textit{de novo} sequencing methods is to use a gold standard produced via database search.
+In this approach, mass spectrometry data derived from a single species is searched against the reference proteome for that species, yielding a ranked list of peptide-spectrum matches (PSMs).
+Including in the peptide database a collection of reversed or shuffled ``decoy'' peptides provides a rigorous way to set a threshold in this list of PSMs while controlling the false discovery rate (FDR) among the PSMs above the threshold \cite{elias:target}.
+The resulting set of high-confidence PSMs can be used either to train or evaluate a \textit{de novo} sequencing model.
 
-\fixme{Talk about complications.  Labeling strategy.  Train/test leakage.  Batch effects.  Quality of the spectra.}
+Some version of the above protocol has been used to develop labeled training and validation data for essentially every published deep learning \textit{de novo} sequencing method.
+One exception is cases such as \fixme{XXX} that use spectra from synthesized peptide sequences for training.
+However, even in these cases, a gold standard derived from database search is used for evaluation of the method.
+
+Unfortunately, creating a high quality gold standard set of labeled spectra can be tricky.
+One challenge is ensuring that the search strategy employs appropriate parameters.
+For instance, one widely used benchmark dataset \cite{tran2017denovo} used a search strategy that failed to account for missassigned isotopic peaks during the acquisition stage.
+This error led to frequently assigning a deamidation modification, when the observed mass shift was better explained by an isotopic mass shift on the precursor \textit{m/z} \cite{bittremieux:deep}.
+A second challenge relates to the notion of train/test ``leakage,'' in which information used to train the model leaks into the evaluation procedure.
+In the \textit{de novo} setting, a common mistake is to randomly segregate a given set of labeled spectra into training and test sets, without regard to the associated peptides.
+As a result, spectra generated by the same peptide sequence may occur in both the training and test sets.
+Such duplicated peptides give an unfair advantage to the sequencing method, and the leakage will be even more useful to parameter-rich methods that are capable of memorizing many features of the training data.
+
+In this work, we revisit the nine-species benchmark dataset that was employed in the first deep learning \textit{de novo} sequencing method, DeepNovo \cite{trans2017denovo}.
+This is a widely used dataset, which has been employed for training or evaluation in at least \fixme{XXX} subsequent studies \cite{XXX}.
+The setup is quite straightforward.
+The authors downloaded nine publicly available datasets, all of which were generated on a Thermo Scientific Q Exactive mass spectrometer, and each of which was carried out in a different species.
+Each dataset was searched against the corresponding reference proteome, using a target-decoy strategy to accept a set of PSMs subject to a PSM-level FDR threshold of 1\%.
+Because the data are derived from different species, the peptides in each set are largely (but not entirely) disjoint.
+To use the benchmark, it is typical to apply a cross-validation strategy, in which a model is trained on eight species and tested on the held-out species, and the procedure is repeated nine different ways.
+
+In developing our Casanovo \textit{de novo} sequencing model, we identified several problems with the nine-species benchmark \cite{yilmaz2023sequence}.
+These included the deamidation problem mentioned above, as well as some uncertainty regarding how the FDR was controlled.
+Perhaps most importantly, we recognized that a non-negligible proportion of peptides are shared among the different species, with the highest overlap between human and mouse.
+
+In light of these difficulties, we downloaded the same datasets from the PRIDE repository and systematically reanalyzed all of the data, using a standard search procedure---the Tide search engine \cite{diament2011faster} followed by Percolator \cite{kall2007semi-supervised} with PSM-level FDR control at 1\%.
+We then filtered the PSMs to prevent any peptide sequence from appearing in more than one species.
+Finally, because some of the single-species datasets are markedly larger than others, we produced a more balanced version of the dataset.
+Hence, we make publicly available all three versions of this dataset: the original, unfiltered dataset if you simply want to train a model with as much data as possible (``overlap''), the peptide-disjoint dataset that can be used to avoid train/test leakage (``main''), and the reduced peptide-disjoint dataset if you want your analysis to run more quickly (``balanced'').
+In addition, we make available all of the intermediate files, for use in validating the benchmark.
 
@@
 
-We created a new version of the nine-species benchmark originally described by Tran {\em et al.} \cite{tran2017denovo}
+\begin{table}
+\scriptsize
+\centering
+\begin{tabular}{clcrrrr}
+\hline
+PRIDE & Species & Uniprot & Files & Spectra & precursor & fragment \\
+\hline
+PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932848 & 20 & 0.05 \\
+PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306786 10 & 0.05 \\
+PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3728183 & 10 & 0.05 \\
+PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4336428 & 30 & 0.05 \\
+PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2272023 & 20 & 0.05 \\
+PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603506 & 15 & 0.05 \\
+PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1477397 & 20 & 0.05 \\
+PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823169 & 20 & 0.05 \\
+PXD004424 & \textit{H. sapiens} & UP000005640 & 26 & 684821 & 20 & 0.02 \\
+\hline
+Total & & & 343 & 15,165,161 \\
+\end{tabular}
+\caption{{\bf The nine-species benchmark.}
+The final two columns specify the precursor window size (in ppm) and fragment bin size (in Da) used in the database search step.
+No reference proteome is available for \textit{Vigna mungo}, so the proteome for the closely related species \textit{Vigna radiata} was used instead.
+}
+\label{tab:benchmark}
+\end{table}
 
+For our benchmark, we used the same nine studies originally identified by Tran {\em et al.} \cite{tran2017denovo}
 \begin{enumerate}
@@
 
-To do so, we downloaded the RAW files from the same nine PRIDE projects (Supplementary Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4.
-We also downloaded the corresponding nine Uniprot reference proteomes and constructed a Tide index for each one, using Crux version 4.1.
-For one species (\textit{Vigna mungo}), no reference proteome is available, so we used the proteome of the closely related species \textit{Vigna radiata}.
+We downloaded the RAW files from the corresponding PRIDE projects (Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4.
+We downloaded the corresponding nine Uniprot reference proteomes and constructed a Tide index for each one, using Crux version 4.2.
+Note that, for one species (\textit{Vigna mungo}) no reference proteome is available, so we used the proteome of the closely related species \textit{Vigna radiata}.
 
-\subsection*{Assigning peptide labels to spectra}
+\begin{table}
+\scriptsize
+\centering
+\begin{tabular}{lrrrrrr}
+\hline
+Species & \multicolumn{2}{c}{Overlap} & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} \\
+& PSMs & Peptides  & PSMs & Peptides  & PSMs & Peptides  \\
+\hline
+\textit{Vigna mungo} & & & 108514 & 12001 \\
+\textit{Mus musculus} & & & 25541 & 5899 \\
+\textit{Methanosarcina mazei} & & & 267333 & 15925 \\
+\textit{Bacillus subtilis} & & & 1358337 & 30786 \\
+\textit{Candidatus endoloripes} & & & 82290 & 8392 \\
+\textit{Solanum lycopersicum} & & & 178413 & 49745 \\
+\textit{Saccharomyces-cerevisiae} & & & 585593 & 19720 \\
+\textit{Apis mellifera} & & & 194281 & 21559 \\
+\textit{H. sapiens} & & & 44604 & 11289 \\
+\hline
+Total & & & 2,844,906 & 175,316 \\
+\end{tabular}
+\caption{{\bf Three versions of the nine-species benchmark.}
+  The ``overlap'' dataset contains peptides that are associated with more than one species.
+  The ``main'' dataset \fixme{XXX}
+}
+\label{tab:benchmark}
+\end{table}
 
-We specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
+\subsection*{Database search and FDR control}
+
+We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
+In creating the Tide index, we specified specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
 Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
-Each index also contains a shuffled decoy peptide corresponding to each target peptide.
-Each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Supplementary Table~\ref{tab:benchmark}).
-We used XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
+Tide automatically added to each index a shuffled decoy peptide corresponding to each target peptide.
+Thereafter, each MGF file was searched against the corresponding index using the precursor window size and fragment bin tolerance specified in the original study (Table~\ref{tab:benchmark}).
+The search engine employed XCorr scoring with Tailor calibration \cite{sulimov2020tailor}, and we allowed for 1 isotope error in the selection of candidate peptides.
 All search results were then analyzed jointly per species using the Crux implementation of Percolator, with default parameters.
@@
 We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
+The resulting benchmark (``overlap'') contains \fixme{XXX} million PSMs drawn from 343 RAW files.
 
@@
 
-We then post-processed the PSMs to eliminate peptides that are shared between species.
+To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
 Among the 229,984 unique peptides, we identified 3797 (1.7\%) that occur in more than one species.
 For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
-Note that when identifying shared peptides between species, we
-considered all modified forms of a given peptide sequence to be the
-same.  Hence, if a given peptide appears in more than one species,
-then that peptide, including all its modified forms, is randomly
-assigned to a single species and eliminated from the others.
-The final benchmark dataset consists of 2.8 million PSMs drawn from 343 RAW files.
+Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same.
+Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
+The final, non-redundant benchmark dataset (``main'') consists of 2.8 million PSMs corresponding to \fixme{XXX} peptides.
 
@@
 
-At this stage, the benchmark is quite imbalanced, in the sense that some species have a much larger number of associated PSMs.
+At this stage, the benchmark was quite imbalanced, in the sense that some species had a much larger number of associated PSMs.
 We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
@@
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset consists of \fixme{XXX} PSMs.
+The final, balanced benchmark dataset consists of \fixme{XXX} PSMs from \fixme{XXX} peptides.
 
@@
 
-\section*{Usage Notes (optional)}
-
 \section*{Code Availability}
 
-\section*{References}
+\bibliographystyle{unsrt}
+\bibliography{refs} % See https://github.com/Noble-Lab/noble-lab-references
 
@@
 \paragraph{Competing Interests}
-
-\bibliographystyle{unsrt}
-\bibliography{refs} % See https://github.com/Noble-Lab/noble-lab-references
-
-\section*{Supplementary information}
-
-\appendix
-\renewcommand{\theequation}{S\arabic{equation}}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\renewcommand{\thesection}{S\arabic{section}}
-\renewcommand{\thetable}{S\arabic{table}}
-\setcounter{table}{0}
-\setcounter{figure}{0}
+These authors declare that they have no competing interests.
 
 \end{document}
-
-% How to make a dense list.
-%\usepackage{enumitem}
-%\begin{enumerate}[noitemsep,nolistsep]

### 2024-07-15 bin/summarize_benchmark.py
Various changes

@@ -23,8 +23,10 @@ def count_spectra(mgf_filenames):
     return return_value
 
 def count_lines(filename):
-    with open(filename, "r") as my_file:
-        return(len(my_file.readlines()))
+    if os.path.exists(filename):
+        with open(filename, "r") as my_file:
+            return(len(my_file.readlines()))
+    return 0
 
 ###########################################################################
 # MAIN
@@ -49,52 +51,48 @@ def main():
     driver = pandas.read_csv(args.driver_filename, sep="\t", header=None)
     all_species = list(driver[1])
 
-    column_names = ["#raw", "#mgf", "#spectra", "#PSMs", "#peptides"]
     output = {} # Key = column name, value = list of entries.
     output["species"] = all_species
+    column_names = ["#raw", "#mgf", "#spectra", "#PSMs", "#peptides"]
     for stat in column_names:
         output[stat] = []
 
     # Gather up all the statistics
     for species in all_species:
-        for stat in column_names:
-            if stat == "#raw":
-                output[stat].append(
-                    len(glob.glob(os.path.join(args.data_dir,
-                                               species, "*.mgf")))
-                )
-            elif stat == "#mgf":
-                output[stat].append(
-                    len(glob.glob(os.path.join(args.benchmark_dir,
-                                               species, "*.mgf")))
-                )
-            elif stat == "#spectra":
-                output[stat].append(
-                    count_spectra(glob.glob(os.path.join(args.data_dir,
-                                                         species, "*.mgf")))
-                )
-            elif stat == "#PSMs":
-                output[stat].append(
-                    count_spectra(glob.glob(os.path.join(args.benchmark_dir,
-                                                         species, "*.mgf")))
-                )
-            elif stat == "#peptides":
-                peptide_filename = os.path.join(args.benchmark_dir,
-                                                species, "peptides.txt")
-                try:
-                    num_peptides = count_lines(peptide_filename)
-                except:
-                    num_peptides = 0
-                output[stat].append(num_peptides)
+        print(species, file=sys.stderr)
+        output["#raw"].append(
+            len(glob.glob(os.path.join(args.data_dir, species, "*.mgf")))
+        )
+        output["#mgf"].append(
+            len(glob.glob(os.path.join(args.benchmark_dir, species, "*.mgf")))
+        )
+        output["#spectra"].append(
+            count_spectra(glob.glob(os.path.join(args.data_dir,
[…]

### 2024-07-15 bin/summarize_benchmark.py
Update script

@@ -26,6 +26,26 @@ def count_lines(filename):
     with open(filename, "r") as my_file:
         return(len(my_file.readlines()))
 
+def clean_peptide (peptide):
+    "Remove PTMs."
+
+    return re.sub(r"[0-9\.\[\]]+", "", peptide)
+    
+
+def count_peptides(mgf_filenames):
+    """Count number of distinct peptides in a set of MGF files.  Eliminates
+    modifications first."""
+
+    peptides = {} # Key = peptide, value = True
+    for mgf_filename in mgf_filenames:
+        with open(mgf_filename, "r") as mgf_file:
+            for line in mgf_file:
+                if line.startswith("SEQ="):
+                    peptides[clean_peptide(line.rstrip().split("=")[1])] = True
+
+    return len(peptides)
+                            
+    
 ###########################################################################
 # MAIN
 ###########################################################################
@@ -58,6 +78,9 @@ def main():
     # Gather up all the statistics
     for species in all_species:
         for stat in column_names:
+
+            mgf_list = 
+            
             if stat == "#raw":
                 output[stat].append(
                     len(glob.glob(os.path.join(args.data_dir,
@@ -84,7 +107,10 @@ def main():
                 try:
                     num_peptides = count_lines(peptide_filename)
                 except:
-                    num_peptides = 0
+                    num_peptides = count_peptides(
+                        glob.glob(os.path.join(args.benchmark_dir,
+                                               species, "*.mgf"))
+                    )
                 output[stat].append(num_peptides)
 
     output["precursor"] = list(driver[2])

### 2024-07-16 bin/summarize_benchmark.py
Edit script

@@ -80,25 +80,16 @@ def main():
     # Gather up all the statistics
     for species in all_species:
         print(species, file=sys.stderr)
-        output["#raw"].append(
-            len(glob.glob(os.path.join(args.data_dir, species, "*.mgf")))
-        )
-        output["#mgf"].append(
-            len(glob.glob(os.path.join(args.benchmark_dir, species, "*.mgf")))
-        )
-        output["#spectra"].append(
-            count_spectra(glob.glob(os.path.join(args.data_dir,
-                                                 species, "*.mgf")))
-        )
-        output["#PSMs"].append(
-            count_spectra(glob.glob(os.path.join(args.benchmark_dir,
-                                                 species, "*.mgf")))
-        )
-        output["#peptides"].append(
-            count_peptides(
-            count_lines(os.path.join(args.benchmark_dir,
-                                     species, "peptides.txt"))
-        )
+
+        data_mgfs = glob.glob(os.path.join(args.data_dir, species, "*.mgf"))
+        benchmark_mgfs = glob.glob(os.path.join(args.benchmark_dir, species,
+                                                "*.mgf"))
+        
+        output["#raw"].append(len(data_mgfs))
+        output["#mgf"].append(len(benchmark_mgfs))
+        output["#spectra"].append(count_spectra(data_mgfs))
+        output["#PSMs"].append(count_spectra(benchmark_mgfs))
+        output["#peptides"].append(count_peptides(benchmark_mgfs))
 
     output["precursor"] = list(driver[2])
     output["fragment"] = list(driver[3])

### 2024-07-16 bin/summarize_benchmark.py
Fix bugs in scripts

@@ -6,6 +6,7 @@ import argparse
 import pandas
 import os
 import glob
+import re
 
 DESCRIPTION = """Write text and HTML tables describing a given
 benchmark dataset."""
@@ -82,8 +83,9 @@ def main():
         print(species, file=sys.stderr)
 
         data_mgfs = glob.glob(os.path.join(args.data_dir, species, "*.mgf"))
-        benchmark_mgfs = glob.glob(os.path.join(args.benchmark_dir, species,
-                                                "*.mgf"))
+        benchmark_mgfs = glob.glob(
+            os.path.join(args.benchmark_dir, species, "*.mgf")
+        )
         
         output["#raw"].append(len(data_mgfs))
         output["#mgf"].append(len(benchmark_mgfs))

### 2024-07-16 bin/annotate_mgf.py
Add new scripts

@@ -1,0 +1,134 @@
+#!/usr/bin/env python
+# AUTHOR: William Stafford Noble
+# CREATE DATE: 30 Aug 2022
+import sys
+
+USAGE = """USAGE: annotate_mgf.py <threshold> <log> <percolator> <mgf>
+
+Use Percolator PSM-level output to annotate a given MGF file,
+including only PSMs with q-value better than the specified threshold.
+
+Inputs are
+ o the desired FDR threshold,
+ o the log file produced by Crux percolator,
+ o the PSM-level Percolator output file, and
+ o the MGF file to be annotated.
+
+The resulting annotated MGF is printed to stdout.
+
+Note that spectra with no assigned charge state are skipped.
+
+Searches the log file for the assigned file index in a line like this:
+
+INFO: Assigning index <integer> to <file>.
+
+Assumes that the Percolator file has PSM IDs of the form
+<foo>_<file-index>_<scan-number>
+
+"""
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+    global USAGE
+
+    # Parse the command line.
+    if (len(sys.argv) != 5):
+        sys.stderr.write(USAGE)
+        sys.exit(1)
+    fdr_threshold = float(sys.argv[1])
+    log_filename = sys.argv[2]
+    percolator_filename = sys.argv[3]
+    mgf_filename = sys.argv[4]
+
+    # Get the file index from the log file.
+    target_file_index = None
+    with open(log_filename, "r") as log_file:
+        for line in log_file:
+            words = line.rstrip().split()
+            if ( (len(words) == 6) and
+                 (words[0] == "INFO:") and
+                 (words[1] == "Assigning") and
+                 (words[2] == "index") and
+                 (words[5][:-1] == mgf_filename) ): # Remove trailing period.
+                target_file_index = int(words[3])
+                break
+    if (target_file_index == None):
+        print(f"Cannot find {mgf_filename} in {log_filename}.", file=sys.stderr)
+        sys.exit(1)
+    print(f"File index: {target_file_index}.", file=sys.stderr)
+
+    # Read the Percolator PSMs into a dictionary.
+    psms = {} # Key = scan number, value = peptide
+    with open(percolator_filename, "r") as percolator_file:
+
+        header = percolator_file.readline().rstrip().split("\t")
+        psm_id_col = header.index("PSMId")
+        qvalue_col = header.index("q-value")
+        peptide_col = header.index("peptide")
+
+        for line in percolator_file:
+            words = line.rstrip().split("\t")
[…]

### 2024-07-16 bin/clean-benchmark.py

@@ -1,0 +1,172 @@
+#!/usr/bin/env python
+# AUTHOR: William Stafford Noble
+# CREATE DATE: 23 Sep 2022
+import sys
+import argparse
+import os
+import random
+import re
+from pyteomics import mgf
+
+DESCRIPTION = """
+Clean up the 9-species benchmark to eliminate peptides that are shared
+between species.  The old root directory is assumed to contain one
+subdirectory per species, with annotated MGFs in each subdirectory.
+The script creates a copy of that directory structure in the specified
+new location.  For any peptide that appears in more than one species,
+one species is selected at random, and all PSMs associated with that
+peptide in other species are removed.
+
+At the same time, convert PTMs from Tide format to Casanovo format.
+"""
+
+def convert_ptms (peptide):
+    "Convert PTMs in a peptide from Tide to Casanovo format."
+
+    # Handle internal modifications.
+    mod_dict = {"C":"C+57.021",
+                "M[15.9949]":"M+15.995",
+                "[0.9840]":"+0.984"}
+    for dn_mod in mod_dict.keys():
+        peptide = peptide.replace(dn_mod, mod_dict[dn_mod])
+
+    # Handle terminal modifications.
+    n_term_mods = { "[-17.0265]":"-17.027",
+                    "[42.0106]":"+42.011",
+                    "[43.0058]":"+43.006",
+                    "[25.9803]":"+43.006-17.027"}
+    for n_mod in n_term_mods.keys():
+        if n_mod in peptide:
+            peptide = n_term_mods[n_mod]+peptide.replace(n_mod, "")
+
+    return peptide
+
+def i2l(peptide):
+    "Leucines to isoleucines."
+    return re.sub("I", "L", peptide)
+
+def clean_peptide (peptide, do_i2l):
+    "Remove flanking amino acids and PTMs."
+
+    no_mods = re.sub(r"[0-9\.\[\]]+", "", peptide)
+
+    if do_i2l:
+        return i2l(no_mods)
+    return no_mods
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+    global DESCRIPTION
+
+    random.seed(7718)
+
+    # Parse the command line.
+    parser = argparse.ArgumentParser(description=DESCRIPTION)
+    parser.add_argument('--old_root', type=str, required=True,
+                        help="Directory of old benchmark")
+    parser.add_argument('--new_root', type=str, required=True,
+                        help="Directory of new benchmark")
[…]

### 2024-07-17 doc/nine-species/nine-species.tex
Small updates

@@
 \hline
-Species & \multicolumn{2}{c}{Overlap} & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} \\
-& PSMs & Peptides  & PSMs & Peptides  & PSMs & Peptides  \\
+Species & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} \\
+& PSMs & Peptides  & PSMs & Peptides  \\
 \hline
@@
 \end{tabular}
-\caption{{\bf Three versions of the nine-species benchmark.}
-  The ``overlap'' dataset contains peptides that are associated with more than one species.
-  The ``main'' dataset \fixme{XXX}
+\caption{{\bf Two versions of the nine-species benchmark.}
+  The balanced dataset eliminates some data in order to approximately balance the amount of data associated with each species.
 }
@@
 We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
-The resulting benchmark (``overlap'') contains \fixme{XXX} million PSMs drawn from 343 RAW files.
+At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 179,256 distinct peptides.
 
@@
 For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
-Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same.
+Note that when identifying shared peptides between species, we considered all modified forms of a given peptide sequence to be the same, and we converted all isoleucines to leucines.
 Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
-The final, non-redundant benchmark dataset (``main'') consists of 2.8 million PSMs corresponding to \fixme{XXX} peptides.
+The final, non-redundant benchmark dataset (``main'') consists of 2,843,631 PSMs corresponding to 258,141 distinct peptides.
 
@@
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset consists of \fixme{XXX} PSMs from \fixme{XXX} peptides.
+The final, balanced benchmark dataset is approximately $\frac{1}{4}$ the size of the main benchmark, consisting of 779,796 PSMs from 180,250 distinct peptides.
 
@@
 \item PSM-level Percolator results files for targets, in tab-delimited format.
-\item Annotated MGF files for all three versions of the benchmark (overlap, main, and balanced).
+\item Annotated MGF files for both versions of the benchmark (main and balanced).
 \end{itemize}

### 2024-07-17 bin/clean-benchmark.py

@@ -94,8 +94,7 @@ def main():
                             species_mapping[peptide].append(species)
         print(f"Found {len(species_mapping)} peptides.", file=sys.stderr)
 
-    # If a peptide appears in more than one species, select one
-    # species randomly.
+    # If a peptide appears in more than one species, select one randomly.
     num_duplicates = 0
     eliminated_psms = 0
     for peptide in species_mapping.keys():
@@ -123,10 +122,10 @@ def main():
               file=sys.stderr)
 
         print(f"Creating cleaned MGFs for {species}.", file=sys.stderr)
-        mgf_list = [ f.name for f in os.scandir(
-            os.path.join(args.old_root, species)
-        )
-                     if f.is_file() ]
+        mgf_list = [
+            f.name for f in os.scandir(os.path.join(args.old_root, species))
+            if f.is_file()
+        ]
         for mgf_file in mgf_list:
             old_mgf_filename = os.path.join(args.old_root, species, mgf_file)
             new_mgf_filename = os.path.join(args.new_root, species, mgf_file)

### 2024-07-17 bin/summarize_benchmark.py

@@ -29,15 +29,19 @@ def count_lines(filename):
             return(len(my_file.readlines()))
     return 0
 
+def i2l(peptide):
+    "Leucines to isoleucines."
+    return re.sub("I", "L", peptide)
+
 def clean_peptide (peptide):
     "Remove PTMs."
 
-    return re.sub(r"[0-9\.\[\]]+", "", peptide)
+    return i2l(re.sub(r"[0-9\.\[\]]+", "", peptide))
     
 
 def count_peptides(mgf_filenames):
     """Count number of distinct peptides in a set of MGF files.  Eliminates
-    modifications first."""
+    modifications and converts isoleucines first."""
 
     peptides = {} # Key = peptide, value = True
     for mgf_filename in mgf_filenames:

### 2024-07-17 doc/nine-species/nine-species.tex
Edits to paper

@@
 
-\newcommand{\fixme}[1]{{\color{red}{\bf FIXME: #1}\color{black}}}
+\newcommand{\fixme}[1]{{\color{red}{\bf #1}\color{black}}}
 
@@
 \hline
-PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932848 & 20 & 0.05 \\
-PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306786 10 & 0.05 \\
-PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3728183 & 10 & 0.05 \\
-PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4336428 & 30 & 0.05 \\
-PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2272023 & 20 & 0.05 \\
-PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603506 & 15 & 0.05 \\
-PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1477397 & 20 & 0.05 \\
-PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823169 & 20 & 0.05 \\
-PXD004424 & \textit{H. sapiens} & UP000005640 & 26 & 684821 & 20 & 0.02 \\
+PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 & 20 & 0.05 \\
+PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 10 & 0.05 \\
+PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 & 10 & 0.05 \\
+PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 & 30 & 0.05 \\
+PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 & 20 & 0.05 \\
+PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 & 15 & 0.05 \\
+PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 & 20 & 0.05 \\
+PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 & 20 & 0.05 \\
+PXD004424 & \textit{H. sapiens} & UP000005640 & 26 & 684,821 & 20 & 0.02 \\
 \hline
@@
 
-For our benchmark, we used the same nine studies originally identified by Tran {\em et al.} \cite{tran2017denovo}
+For our benchmark, we used the same nine studies originally identified by Tran {\em et al.}. \cite{tran2017denovo}
 \begin{enumerate}
@@
 \item Nevo \textit{et al.}\ studied a rare autosomal recessive lysosomal storage disorder, cystinosis, by carrying out SILAC proteomic analysis of engineered mouse cell lines that harbor a known pathenogenic mutation of the causative gene, \textit{CTNS} \cite{nevo2017impact}.
-\item Cassidy \textit{et al.}\ evaluated two different analytical approaches for carrying out full proteome analysis while identifying short open reading frames: high/low pH reversed phase LC-MS bottom-up approach and a semi-top-down strategy involving separation of proteins a GelFree system followed by digestion and LC-MS analysis \cite{cassidy2016combination}.
+\item Cassidy \textit{et al.}\ evaluated two different analytical approaches for carrying out full proteome analysis while identifying short open reading frames: a high/low pH reversed phase LC-MS bottom-up approach and a semi-top-down strategy involving separation of proteins in a GelFree system followed by digestion and LC-MS analysis \cite{cassidy2016combination}.
 The experiments were carried out using the methane producing archaeon \textit{Methanosarcina mazei}.
@@
 Proteomic analysis was carried out on mushroom bodies and antennae of adult honeybees with and without VSH.
-\item Cypryck \textit{et al.} characterized extracellular vesicles released from human primary macrophages after infection with influenza A viruses \cite{cypryk2017proteomic}.
+\item Cypryck \textit{et al.}\ characterized extracellular vesicles released from human primary macrophages after infection with influenza A viruses \cite{cypryk2017proteomic}.
 \end{enumerate}
@@
 }
-\label{tab:benchmark}
+\label{tab:sizes}
 \end{table}
@@
 
-\fixme{False discovery rate curves}
+\begin{figure}
+  \centering
+  \begin{tabular}{ll}
+    \fixme{Insert figure} & \fixme{Insert figure} \\
+    a & b \\
+    \fixme{Insert figure} & \fixme{Insert figure} \\
+    c & d \\
+  \end{tabular}
+  \caption{\textbf{Validation of the benchmarks.}
+    \textbf{a.} Each series indicates the number of accepted PSMs at 1\% PSM-level FDR for a given species.
+    \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
+    \textbf{c.} Each violin plot indicates the distribution of the proportion of total ion current that is explained by PSMs per species.
+    \textbf{d.} Each violin plots indicates the distribution of correlation values between observed and predicted fragment ion intensities per species.}
+  \label{fig:fdr}
+\end{figure}
+
+Data quality and interpretability varies dramatically from study to study, due to differences in sample type, sample preparation protocols, chromatography and instrument settings, and database size.
+To assess the overall rate of successful identification of spectra in each data set, we plotted the number of accepted PSMs as a function of PSM-level FDR threshold (Figure~\ref{fig:fdr}a).
+\fixme{Describe plot.}
+We also computed the proportion of spectra that were accepted at 1\% PSM-level FDR per species (Figure~\ref{fig:fdr}b).
+\fixme{Describe plot.}
+
+To assess the validity of the PSMs themselves, we carried out two additional validations.
+First, for each PSM, we computed the proportion of the total ion current that is explained by matched b- and y-ions, using a matching tolerance of 0.05~\textit{m/z}.
+The resulting distributions of proportions differ substantially by species (Figure~\ref{fig:fdr}c).
+\fixme{Describe plot.}
+Second, for each PSM, we used the \fixme{XXX} model to predict b/y-ion peak intensities, and we computed the Pearson correlation between these observed and predicted intensities.
+Again, this distribution of correlations differs substantially by species (Figure~\ref{fig:fdr}d).
+\fixme{Describe plot.}
 
 \section*{Code Availability}
+All code required to generate the various benchmarks and to produce the figures in this manuscript is available with an Apache license at \url{https://github.com/Noble-Lab/nine-species}, with a snapshot of the repository stored at \fixme{zenodo}.
 

### 2024-07-17 doc/nine-species/nine-species.tex
Minor edits

@@
 Recently, as in many other areas of science, considerable progress toward solving this challenge has been made using deep learning, in which multi-layer neural networks with millions of parameters are trained to generate peptide sequences from observed spectra.
-The first such deep learning method, DeepNovo \cite{tran2017denovo}, has been followed by at least 22 additional publications (reviewed in \cite{XXX}).
+The first such deep learning method, DeepNovo \cite{tran2017denovo}, has been followed by at least 22 additional publications (reviewed in \cite{bittremieux2024deep}).
 
@@
 In this approach, mass spectrometry data derived from a single species is searched against the reference proteome for that species, yielding a ranked list of peptide-spectrum matches (PSMs).
-Including in the peptide database a collection of reversed or shuffled ``decoy'' peptides provides a rigorous way to set a threshold in this list of PSMs while controlling the false discovery rate (FDR) among the PSMs above the threshold \cite{elias:target}.
+Including in the peptide database a collection of reversed or shuffled ``decoy'' peptides provides a rigorous way to set a threshold in this list of PSMs while controlling the false discovery rate (FDR) among the PSMs above the threshold \cite{elias2007target}.
 The resulting set of high-confidence PSMs can be used either to train or evaluate a \textit{de novo} sequencing model.
@@
 For instance, one widely used benchmark dataset \cite{tran2017denovo} used a search strategy that failed to account for missassigned isotopic peaks during the acquisition stage.
-This error led to frequently assigning a deamidation modification, when the observed mass shift was better explained by an isotopic mass shift on the precursor \textit{m/z} \cite{bittremieux:deep}.
+This error led to frequently assigning a deamidation modification, when the observed mass shift was better explained by an isotopic mass shift on the precursor \textit{m/z} \cite{bittremieux2024deep}.
 A second challenge relates to the notion of train/test ``leakage,'' in which information used to train the model leaks into the evaluation procedure.
@@
 
-In this work, we revisit the nine-species benchmark dataset that was employed in the first deep learning \textit{de novo} sequencing method, DeepNovo \cite{trans2017denovo}.
+In this work, we revisit the nine-species benchmark dataset that was employed in the first deep learning \textit{de novo} sequencing method, DeepNovo \cite{tran2017denovo}.
 This is a widely used dataset, which has been employed for training or evaluation in at least \fixme{XXX} subsequent studies \cite{XXX}.
@@
 
-In developing our Casanovo \textit{de novo} sequencing model, we identified several problems with the nine-species benchmark \cite{yilmaz2023sequence}.
+In developing our Casanovo \textit{de novo} sequencing model, we identified several problems with the nine-species benchmark \cite{yilmaz2024sequence}.
 These included the deamidation problem mentioned above, as well as some uncertainty regarding how the FDR was controlled.

### 2024-07-17 bin/construct-benchmark.py
Add script from other repo

@@ -1,0 +1,124 @@
+#!/usr/bin/env python
+# AUTHOR: William Stafford Noble
+# CREATE DATE: 30 Aug 2022
+import sys
+import datetime
+import os
+import ppx
+import subprocess
+
+USAGE = """USAGE: construct-benchmark.py <listing> <output dir>
+
+Download the raw files associated with the DeepNovo benchmark dataset
+from Tran et al., PNAS, 2017.  The input listing is a two-column text
+file containing PRIDE IDs and species names.  The raw files are
+downloaded into the specified output directory in subdirectories named
+by species.
+"""
+
+# The location of the Thermo RAW file parser.
+THERMORFP = "/Users/wnoble/opt/miniconda3/bin/thermorawfileparser"
+if not os.path.exists(THERMORFP):
+    THERMORFP = "~/proj/thermorawfileparser/ThermoRawFileParser.exe"
+if not os.path.exists(THERMORFP):
+    THERMORFP = "~/software/ThermoRawFileParser.v1.3.4/ThermoRawFileParser.exe"
+
+#############################################################################
+def run_command(command, outputFileName):
+    """Run a command with error checking."""
+
+    # Skip the command if the output file already exists.
+    if (outputFileName != "") and (os.path.exists(outputFileName)):
+        sys.stderr.write("%s exists.\n" % outputFileName)
+        return
+  
+    sys.stderr.write("RUN: %s\n" % command)
+    try:
+        returnCode = subprocess.call(command, shell=True)
+        if (returnCode != 0):
+            sys.stderr.write("Child was terminated by signal %d\n" 
+                             % -returnCode)
+#            sys.exit(1)
+    except OSError as e:
+        sys.stderr.write("Execution failed: %s\n" % e)
+        sys.exit(1)
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+    global USAGE
+
+    # Parse the command line.
+    if (len(sys.argv) != 3):
+        sys.stderr.write(USAGE)
+        sys.exit(1)
+    listing_filename = sys.argv[1]
+    output_dir = sys.argv[2]
+
+    # Read the PRIDE IDs.
+    prides = {} # Key = species, value = PRIDE identifier
+    with open(listing_filename, "r") as listing_file:
+        for line in listing_file:
+            (pride, species) = line.rstrip().split()[:2]
+            prides[species] = pride
+    print(f"Read {len(prides)} entries from {listing_filename}.", 
+          file=sys.stderr)
+
+    # Make a small README in the output directory.
[…]

### 2024-07-17 bin/match_by.py

@@ -1,0 +1,51 @@
+#!/usr/bin/env python
+import sys
+import argparse
+import matplotlib.pyplot as plt
+import spectrum_utils.spectrum
+import spectrum_utils.plot
+import pyteomics.mgf
+
+DESCRIPTION = """Given an annotated MGF, compute the proportion of intensity
+matched to b- and y-ions.  Produces a tsv file (<root>.txt) and a plot
+(<root>.pdf)."""
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+
+    # Parse the command line.
+    parser = argparse.ArgumentParser(description=DESCRIPTION)
+    parser.add_argument('--mgf', type=str, required=True,
+                        help="Annotated MGF file")
+    parser.add_argument('--root', type=str, required=True,
+                        help="Output file root")
+    args = parser.parse_args()
+
+    title = f"controllerType=0 controllerNumber=1 scan={args.scan}"
+    spectrum = pyteomics.mgf.get_spectrum(args.mgf, title)
+
+    # Convert from pyteomics to spectrum_utils format.
+    spectrum = spectrum_utils.spectrum.MsmsSpectrum(
+        title,
+        float(spectrum["params"]["pepmass"][0]), 
+        int(spectrum["params"]["charge"][0]), 
+        spectrum["m/z array"], 
+       spectrum["intensity array"]
+     )
+
+    # Annotate the spectrum.
+    spectrum = (
+        spectrum.set_mz_range(min_mz=100, max_mz=1400)
+        .remove_precursor_peak(args.tolerance, "ppm")
+        .filter_intensity(min_intensity=0.05, max_num_peaks=50)
+        .scale_intensity("root")
+        .annotate_proforma(
+            args.peptide, args.tolerance, "ppm", ion_types="aby"
+        )
+    )
+
+
+if __name__ == "__main__":
+    main()

### 2024-07-17 bin/match_by.py
Add new script

@@ -7,8 +7,45 @@ import spectrum_utils.plot
 import pyteomics.mgf
 
 DESCRIPTION = """Given an annotated MGF, compute the proportion of intensity
-matched to b- and y-ions.  Produces a tsv file (<root>.txt) and a plot
-(<root>.pdf)."""
+matched to b- and y-ions.  Prints a list of values to stdout."""
+
+def get_percent_matched(spectrum):
+    
+    # Extract the peptide sequence.
+    peptide = spectrum['params']['seq']
+
+    # Convert from pyteomics to spectrum_utils format.
+    spectrum = spectrum_utils.spectrum.MsmsSpectrum(
+        spectrum['params']['title'],
+        float(spectrum['params']['pepmass'][0]), 
+        int(spectrum['params']['charge'][0]), 
+        spectrum['m/z array'], 
+        spectrum['intensity array']
+    )
+    
+    # Annotate the spectrum.
+    spectrum = (
+        spectrum.set_mz_range(min_mz=100, max_mz=1400)
+        .remove_precursor_peak(0.05, "Da")
+        .filter_intensity(min_intensity=0.05, max_num_peaks=50)
+        .scale_intensity("root")
+        .annotate_proforma(peptide, 0.05, "Da", ion_types="by")
+    )
+
+    # Compute percent matched intensity.
+    total_intensity = 0.0
+    matched_intensity = 0.0
+    for annotation, intensity in zip(
+            spectrum.annotation, spectrum.intensity
+    ):
+        total_intensity += intensity
+        if len(annotation.fragment_annotations) > 0:
+#            print(f"Matching {annotation} with intensity {intensity}.")
+            matched_intensity += intensity
+#        else:
+#            print(f"No matching for intensity {intensity}.")
+
+    return 100 * matched_intensity / total_intensity
 
 ###########################################################################
 # MAIN
@@ -19,33 +56,122 @@ def main():
     parser = argparse.ArgumentParser(description=DESCRIPTION)
     parser.add_argument('--mgf', type=str, required=True,
                         help="Annotated MGF file")
-    parser.add_argument('--root', type=str, required=True,
-                        help="Output file root")
     args = parser.parse_args()
 
[…]

### 2024-07-17 bin/match_by.py
Working version of matching script

@@ -6,14 +6,34 @@ import spectrum_utils.spectrum
 import spectrum_utils.plot
 import pyteomics.mgf
 
-DESCRIPTION = """Given an annotated MGF, compute the proportion of intensity
-matched to b- and y-ions.  Prints a list of values to stdout."""
+USAGE = """USAGE: match_by.py <mgf>+
+
+Given one or more annotated MGFs, compute the proportion of intensity
+matched to b- and y-ions in each spectrum. Prints a list of values to
+stdout."""
+
+def convert_ptms (peptide):
+    "Add brackets around PTMs in a peptide."
+
+    mod_dict = {
+        "C+57.021":"C[+57.021]",
+        "M+15.995":"M[+15.9949]",
+        "+0.984":"[+0.9840]",
+        "+42.011":"[+42.011]-",
+        "+43.006-17.027":"[+25.980]-",
+        "+43.006":"[+43.006]-",
+        "-17.027":"[-17.027]-"
+    }
+    for dn_mod in mod_dict.keys():
+        peptide = peptide.replace(dn_mod, mod_dict[dn_mod])
+
+    return peptide
 
 def get_percent_matched(spectrum):
     
     # Extract the peptide sequence.
-    peptide = spectrum['params']['seq']
-
+    peptide = convert_ptms(spectrum['params']['seq'])
+    
     # Convert from pyteomics to spectrum_utils format.
     spectrum = spectrum_utils.spectrum.MsmsSpectrum(
         spectrum['params']['title'],
@@ -35,32 +55,33 @@ def get_percent_matched(spectrum):
     # Compute percent matched intensity.
     total_intensity = 0.0
     matched_intensity = 0.0
-    for annotation, intensity in zip(
-            spectrum.annotation, spectrum.intensity
-    ):
+    for annotation, intensity in zip(spectrum.annotation, spectrum.intensity):
         total_intensity += intensity
         if len(annotation.fragment_annotations) > 0:
-#            print(f"Matching {annotation} with intensity {intensity}.")
             matched_intensity += intensity
-#        else:
-#            print(f"No matching for intensity {intensity}.")
 
-    return 100 * matched_intensity / total_intensity
+    return (peptide, 100 * matched_intensity / total_intensity)
 
 ###########################################################################
 # MAIN
 ###########################################################################
 def main():
+    global USAGE
 
     # Parse the command line.
[…]

### 2024-07-17 results/wnoble/2024-07-17validations/make_histograms.py
Add make_histograms.py

@@ -1,0 +1,32 @@
+#!/usr/bin/env python
+# AUTHOR: WSN
+# CREATE DATE: 17 July 2024
+import sys
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+
+    # Read all the match percentages from species-specific files.
+    match_df = {} # Key = species, value = list of match percentages.
+    for match_filename in sys.argv[1:]:
+        species = match_filename.split(".")[1] # match_by.<species>.txt
+        with open(match_filename, "r") as match_file:
+            match_file.readline()
+            match_df[species] = [line.rstrip().split("\t")[2]
+                                 for line in match_file]
+            print(f"Read {len(match_df[species])} match percentages from " +
+                  f"{species}.", file=sys.stderr)
+            
+
+    # Make the histogram.
+    fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)
+    i = 0
+    for species in match_df.keys():
+        axs[i].hist(match_df[species], bins=100)
+        i += 1
+    plt.show()
+
+if __name__ == "__main__":
+    main()

### 2024-07-18 bin/match_by.py
Fix bug in matching script.

@@ -46,7 +46,7 @@ def get_percent_matched(spectrum):
     # Annotate the spectrum.
     spectrum = (
         spectrum.set_mz_range(min_mz=100, max_mz=1400)
-        .remove_precursor_peak(0.05, "Da")
+        .remove_precursor_peak(2.0, "Da")
         .filter_intensity(min_intensity=0.05, max_num_peaks=50)
         .scale_intensity("root")
         .annotate_proforma(peptide, 0.05, "Da", ion_types="by")
@@ -178,10 +178,235 @@ SEQ=VVQEQGTHPK
 END IONS
 """
 
+big_mgf = """BEGIN IONS
+TITLE=controllerType=0 controllerNumber=1 scan=3919
+PEPMASS=828.3374
+RTINSECONDS=829.10304
+CHARGE=3+
+SCANS=3919
+SEQ=C+57.021ANFDNQDNNHYNHNHNQAR
+120.080719 11572.1962890625 
+126.0542755 2370.3344726563 
+129.0657501 2044.9259033203 
+129.1024933 43609.078125 
+130.0859833 4392.8032226563 
+133.0430756 21430.109375 
+134.0460052 1623.1136474609 
+134.3571014 1661.8100585938 
+141.0658417 11187.9462890625 
+143.1177673 4205.8627929688 
+147.1126099 12227.9482421875 
+155.081955 3122.3088378906 
+156.0751495 2471.1391601563 
+158.0921631 3096.1274414063 
+159.059082 3108.6096191406 
+160.0423584 5466.2817382813 
+166.0613556 3596.2768554688 
+173.1286163 2503.1989746094 
+175.1190186 29479.962890625 
+187.053421 12431.7021484375 
+197.128067 3531.0043945313 
+198.8144684 2500.5632324219 
+204.0802002 60947.6171875 
+205.0826416 2895.0197753906 
+209.1024475 2274.1833496094 
+215.0477295 3179.7243652344 
+215.1389313 2373.5729980469 
+217.099411 2334.8037109375 
+221.1032867 4907.732421875 
+231.0918884 2550.3801269531 
+232.0746613 35466.87109375 
+232.0926819 2682.2934570313 
+234.1236725 2912.5451660156 
+234.1440125 3220.1662597656 
+235.0829468 2804.4089355469 
+235.118576 5699.9150390625 
+240.1328888 2645.8654785156 
+244.1643219 4398.8686523438 
[…]

### 2024-07-18 results/wnoble/2024-07-17validations/make_histograms.py

@@ -2,20 +2,27 @@
 # AUTHOR: WSN
 # CREATE DATE: 17 July 2024
 import sys
+import matplotlib.pyplot as plt
 
 ###########################################################################
 # MAIN
 ###########################################################################
 def main():
 
+    # Check the command line.
+    if len(sys.argv) < 3:
+        print("USAGE: make_histograms <output plot> <mgf>+", file=sys.stderr)
+        sys.exit(1)
+    
     # Read all the match percentages from species-specific files.
     match_df = {} # Key = species, value = list of match percentages.
-    for match_filename in sys.argv[1:]:
+    for match_filename in sys.argv[2:]:
         species = match_filename.split(".")[1] # match_by.<species>.txt
         with open(match_filename, "r") as match_file:
             match_file.readline()
-            match_df[species] = [line.rstrip().split("\t")[2]
+            match_df[species] = [float(line.rstrip().split("\t")[2])
                                  for line in match_file]
+#            match_df[species] = [float(line.rstrip()) for line in match_file]
             print(f"Read {len(match_df[species])} match percentages from " +
                   f"{species}.", file=sys.stderr)
             
@@ -24,9 +31,12 @@ def main():
     fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)
     i = 0
     for species in match_df.keys():
-        axs[i].hist(match_df[species], bins=100)
+        axs[i].hist(match_df[species], density=True, bins=100)
+        axs[i].set_title(species, loc='left')
+        axs[i].spines['top'].set_visible(False)
+        axs[i].spines['right'].set_visible(False)
         i += 1
-    plt.show()
+    plt.savefig(sys.argv[1])
 
 if __name__ == "__main__":
     main()

### 2024-07-19 doc/nine-species/nine-species.tex
Various updates

@@
 This error led to frequently assigning a deamidation modification, when the observed mass shift was better explained by an isotopic mass shift on the precursor \textit{m/z} \cite{bittremieux2024deep}.
-A second challenge relates to the notion of train/test ``leakage,'' in which information used to train the model leaks into the evaluation procedure.
+A second challenge relates to the notion of train/test leakage, in which information used to train the model leaks into the evaluation procedure.
 In the \textit{de novo} setting, a common mistake is to randomly segregate a given set of labeled spectra into training and test sets, without regard to the associated peptides.
@@
 We then filtered the PSMs to prevent any peptide sequence from appearing in more than one species.
+The resulting data set was used to evaluate Casanovo \cite{yilmaz2024sequence}.
 Finally, because some of the single-species datasets are markedly larger than others, we produced a more balanced version of the dataset.
-Hence, we make publicly available all three versions of this dataset: the original, unfiltered dataset if you simply want to train a model with as much data as possible (``overlap''), the peptide-disjoint dataset that can be used to avoid train/test leakage (``main''), and the reduced peptide-disjoint dataset if you want your analysis to run more quickly (``balanced'').
+Hence, we make publicly available both versions of this dataset: the peptide-disjoint dataset that can be used to avoid train/test leakage (``main''), and the reduced peptide-disjoint dataset if you want your analysis to run more quickly (``balanced'').
 In addition, we make available all of the intermediate files, for use in validating the benchmark.
@@
 
+%% Species & #raw & #mgf & #spectra & #PSMs & #peptides & #PSMs & #peptides & precursor & fragment
+%% \hline
+%% \textit{Vigna mungo} & 24 & 23 & 932848 & 108756 & 13768 & 102576 & 13656 & 20 & 0.05
+%% \textit{Mus musculus} & 13 & 9 & 306786 & 25490 & 6833 & 25490 & 6833 & 10 & 0.05
+%% \textit{Methanosarcina mazei} & 72 & 26 & 3728183 & 267269 & 23448 & 100517 & 16774 & 10 & 0.05
+%% \textit{Bacillus subtilis} & 106 & 9 & 4336428 & 1357908 & 64025 & 113613 & 29245 & 30 & 0.05
+%% \textit{Candidatus endoloripes} & 11 & 11 & 2272023 & 82204 & 10863 & 82204 & 10863 & 20 & 0.05
+%% \textit{Solanum lycopersicum} & 60 & 33 & 603506 & 177941 & 62294 & 100325 & 44052 & 15 & 0.05
+%% \textit{Saccharomyces cerevisiae} & 27 & 5 & 1477397 & 586278 & 33008 & 109023 & 20057 & 20 & 0.05
+%% \textit{Apis mellifera} & 17 & 8 & 823169 & 193437 & 30013 & 101700 & 24905 & 20 & 0.05
+%% \textit{H. sapiens} & 26 & 26 & 684821 & 44348 & 13853 & 44348 & 13853 & 20 & 0.02
+%% Total & 356 & 150 & 15165161 & 2843631 & 258105 & 779796 & 180238 
+%% \hline
+
+%% \hline
+%% Species & #raw & #mgf & #spectra & #PSMs & #peptides & #PSMs & #peptides & precursor & fragment
+%% \hline
+%% \textit{Vigna mungo} & 24 & 23 & 932848 
+%% \textit{Mus musculus} & 13 & 9 & 306786 
+%% \textit{Methanosarcina mazei} & 72 & 26 &
+%% \textit{Bacillus subtilis} & 106 & 9 &
+%% \textit{Candidatus endoloripes} & 11 & 11 & 2272023 & 
+%% \textit{Solanum lycopersicum} & 60 & 33 & 603506 & 
+%% \textit{Saccharomyces cerevisiae} & 27 & 5 & 1477397 & 
+%% \textit{Apis mellifera} & 17 & 8 & 823169 & 
+%% \textit{H. sapiens} & 26 & 26 & 684821 & 
+%% Total & 356 & 150 & 15165161 & 
+%% \hline
+
 \begin{table}
+  \centering
   \scriptsize
-\centering
-\begin{tabular}{clcrrrr}
+\begin{tabular}{lllrrrrrrrrrr}
 \hline
-PRIDE & Species & Uniprot & Files & Spectra & precursor & fragment \\
+PRIDE & Species & Uniprot & Files & Spectra &
+\multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} & Pre & Frag \\
+& & & & & PSM & Pep  & PSM & Pep  \\
 \hline
-PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 & 20 & 0.05 \\
-PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 10 & 0.05 \\
-PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 & 10 & 0.05 \\
-PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 & 30 & 0.05 \\
-PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 & 20 & 0.05 \\
-PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 & 15 & 0.05 \\
-PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 & 20 & 0.05 \\
-PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 & 20 & 0.05 \\
-PXD004424 & \textit{H. sapiens} & UP000005640 & 26 & 684,821 & 20 & 0.02 \\
+PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
+108,756 & 13,768 & 102,576 & 13,656 & 20 & 0.05 \\
+PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
+25,490 & 6833 & 25,490 & 6833 & 10 & 0.05 \\
+PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
+267,269 & 23,448 & 100,517 & 16,774 & 10 & 0.05 \\
+ PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
+1,357,908 & 64,025 & 113,613 & 29,245 & 30 & 0.05 \\
+PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
+82,204 & 10,863 & 82,204 & 10,863 & 20 & 0.05 \\
+PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
+177,941 & 62,294 & 100,325 & 44,052 & 15 & 0.05 \\
+PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
+586,278 & 33,008 & 109,023 & 20,057 & 20 & 0.05 \\
+PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
+193,437 & 30,013 & 101,700 & 24,905 & 20 & 0.05 \\
+PXD004424 & \textit{H. sapiens} & UP000005640 & 26 & 684,821 &
+44,348 & 13,853 & 44,348 & 13,853 & 20 & 0.02 \\
 \hline
-Total & & & 343 & 15,165,161 \\
+Total & & & 343 & 15,165,161  & 2,843,631 & 258,105 & 779,796 & 180,238  \\
 \end{tabular}
-\caption{{\bf The nine-species benchmark.}
+\caption{{\bf Two versions of the nine-species benchmark.}
 The final two columns specify the precursor window size (in ppm) and fragment bin size (in Da) used in the database search step.
-No reference proteome is available for \textit{Vigna mungo}, so the proteome for the closely related species \textit{Vigna radiata} was used instead.
 }
@@
 
-For our benchmark, we used the same nine studies originally identified by Tran {\em et al.}. \cite{tran2017denovo}
+For our benchmark, we used the same nine studies originally identified by Tran {\em et al.} \cite{tran2017denovo}.
 \begin{enumerate}
@@
 
-\begin{table}
-\scriptsize
-\centering
-\begin{tabular}{lrrrrrr}
-\hline
-Species & \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} \\
-& PSMs & Peptides  & PSMs & Peptides  \\
-\hline
-\textit{Vigna mungo} & & & 108514 & 12001 \\
-\textit{Mus musculus} & & & 25541 & 5899 \\
-\textit{Methanosarcina mazei} & & & 267333 & 15925 \\
-\textit{Bacillus subtilis} & & & 1358337 & 30786 \\
-\textit{Candidatus endoloripes} & & & 82290 & 8392 \\
-\textit{Solanum lycopersicum} & & & 178413 & 49745 \\
-\textit{Saccharomyces-cerevisiae} & & & 585593 & 19720 \\
-\textit{Apis mellifera} & & & 194281 & 21559 \\
-\textit{H. sapiens} & & & 44604 & 11289 \\
-\hline
-Total & & & 2,844,906 & 175,316 \\
-\end{tabular}
-\caption{{\bf Two versions of the nine-species benchmark.}
-  The balanced dataset eliminates some data in order to approximately balance the amount of data associated with each species.
-}
-\label{tab:sizes}
-\end{table}
-
 \subsection*{Database search and FDR control}
@@
 Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
-The final, non-redundant benchmark dataset (``main'') consists of 2,843,631 PSMs corresponding to 258,141 distinct peptides.
+The final, non-redundant benchmark dataset (``main'') consists of 2,843,631 PSMs corresponding to 258,105 distinct peptides.
 
@@
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset is approximately $\frac{1}{4}$ the size of the main benchmark, consisting of 779,796 PSMs from 180,250 distinct peptides.
+The final, balanced benchmark dataset is approximately $\frac{1}{4}$ the size of the main benchmark, consisting of 779,796 PSMs from 180,238 distinct peptides.
 
@@
     a & b \\
-    \fixme{Insert figure} & \fixme{Insert figure} \\
-    c & d \\
+    \fixme{Insert figure} \\
+    c \\
   \end{tabular}
@@
     \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
-    \textbf{c.} Each violin plot indicates the distribution of the proportion of total ion current that is explained by PSMs per species.
-    \textbf{d.} Each violin plots indicates the distribution of correlation values between observed and predicted fragment ion intensities per species.}
+    \textbf{c.} Each violin plot indicates the distribution of the proportion of total ion current that is explained by PSMs per species.}
   \label{fig:fdr}
@@
 \fixme{Describe plot.}
-
-To assess the validity of the PSMs themselves, we carried out two additional validations.
-First, for each PSM, we computed the proportion of the total ion current that is explained by matched b- and y-ions, using a matching tolerance of 0.05~\textit{m/z}.
+In addition, for each PSM we computed the proportion of the total peak intensities that is explained by matched b- and y-ions, using a matching tolerance of 0.05~\textit{m/z}.
 The resulting distributions of proportions differ substantially by species (Figure~\ref{fig:fdr}c).
-\fixme{Describe plot.}
-Second, for each PSM, we used the \fixme{XXX} model to predict b/y-ion peak intensities, and we computed the Pearson correlation between these observed and predicted intensities.
-Again, this distribution of correlations differs substantially by species (Figure~\ref{fig:fdr}d).
 \fixme{Describe plot.}
@@
 \paragraph{Competing Interests}
-These authors declare that they have no competing interests.
+The authors declare that they have no competing interests.
 

### 2024-07-19 bin/match_by.py
Minor edit

@@ -45,9 +45,7 @@ def get_percent_matched(spectrum):
     
     # Annotate the spectrum.
     spectrum = (
-        spectrum.set_mz_range(min_mz=100, max_mz=1400)
-        .remove_precursor_peak(2.0, "Da")
-        .filter_intensity(min_intensity=0.05, max_num_peaks=50)
+        spectrum.remove_precursor_peak(2.0, "Da")
         .scale_intensity("root")
         .annotate_proforma(peptide, 0.05, "Da", ion_types="by")
     )

### 2024-07-19 results/wnoble/2024-07-17validations/make_histograms.py

@@ -22,10 +22,8 @@ def main():
             match_file.readline()
             match_df[species] = [float(line.rstrip().split("\t")[2])
                                  for line in match_file]
-#            match_df[species] = [float(line.rstrip()) for line in match_file]
             print(f"Read {len(match_df[species])} match percentages from " +
                   f"{species}.", file=sys.stderr)
-            
 
     # Make the histogram.
     fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)

### 2024-07-19 results/wnoble/2024-07-17validations/make_barchart.py
Add script

@@ -1,0 +1,41 @@
+#!/usr/bin/env python
+# AUTHOR: WSN
+# CREATE DATE: 17 July 2024
+import sys
+import pandas
+import re
+import matplotlib.pyplot as plt
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+
+    # Check the command line.
+    if len(sys.argv) != 3:
+        print("USAGE: make_bar_chart <TSV file> <plot> ", file=sys.stderr)
+        sys.exit(1)
+
+    # Read the statistics into a dataframe.
+    stats_df = pandas.read_csv(sys.argv[1], sep="\t", index_col=0)
+    species = list(stats_df.index)[:-1] # Remove total in last column.
+    print(species, file=sys.stderr)
+
+    # Remove hyphens.
+    species = [re.sub("-", " ", one_species) for one_species in species]
+    print(species, file=sys.stderr)
+
+    num_psms = stats_df[["#PSMs"]].values.tolist()
+    num_spectra = stats_df[["#spectra"]].values.tolist()
+    percents = [100 * (my_tuple[0][0]/my_tuple[1][0])
+                for my_tuple in zip(num_psms, num_spectra)][:-1] # Remove total
+    print(percents, file=sys.stderr)
+
+    # Make the barchart.
+    plt.barh(species, percents)
+    plt.xlabel("Percent identified")
+    plt.tight_layout()
+    plt.savefig(sys.argv[2])
+
+if __name__ == "__main__":
+    main()

### 2024-07-20 doc/nine-species/nine-species.tex
Edits

@@
 
-\fixme{Instructions at
-  \url{https://www.nature.com/sdata/publish/submission-guidelines}.}
+%Instructions at
+%  \url{https://www.nature.com/sdata/publish/submission-guidelines}.}
 
 \begin{abstract}
-  \fixme{Insert abstract here.}
+  Training machine learning models for tasks such as \textit{de novo} sequencing or spectral clustering requires large collections of confidently identified spectra.
+  Here we describe a dataset of 2.8 million high-confidence peptide-spectrum matches derived from nine different species.
+  The dataset is based on a previously described benchmark but has been re-processed to ensure consistent data quality and enforce separation of training and test peptides.
 \end{abstract}
@@
 Some version of the above protocol has been used to develop labeled training and validation data for essentially every published deep learning \textit{de novo} sequencing method.
-One exception is cases such as \fixme{XXX} that use spectra from synthesized peptide sequences for training.
+One exception is methods that use spectra from synthesized peptide sequences for training \cite{karunratanakul2019uncovering, eloff2023denovo, liu2023accurate}.
 However, even in these cases, a gold standard derived from database search is used for evaluation of the method.
@@
 In this work, we revisit the nine-species benchmark dataset that was employed in the first deep learning \textit{de novo} sequencing method, DeepNovo \cite{tran2017denovo}.
-This is a widely used dataset, which has been employed for training or evaluation in at least \fixme{XXX} subsequent studies \cite{XXX}.
+This is a widely used dataset, which has been employed for training or evaluation in at least 15 subsequent studies \cite{yang2024introducing, yang2022dpst, yang2019pnovo, xia2024adanovo, wu2023biatnovo, qiao2021computationally, liu2020residual, lee2024bidirectional, klaproth2024deep, jin2023contranovo, ge2022deps, eloff2023denovo, ebrahimi2024transformer, yilmaz2022denovo, yilmaz2024sequence}.
 The setup is quite straightforward.
@@
 193,437 & 30,013 & 101,700 & 24,905 & 20 & 0.05 \\
-PXD004424 & \textit{H. sapiens} & UP000005640 & 26 & 684,821 &
+PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
 44,348 & 13,853 & 44,348 & 13,853 & 20 & 0.02 \\
@@
 \caption{{\bf Two versions of the nine-species benchmark.}
+  Counts of the number of annotated spectra and distinct peptide sequences are provided for the main and balanced versions of the benchmark.
 The final two columns specify the precursor window size (in ppm) and fragment bin size (in Da) used in the database search step.
@@
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset is approximately $\frac{1}{4}$ the size of the main benchmark, consisting of 779,796 PSMs from 180,238 distinct peptides.
+The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,796 PSMs from 180,238 distinct peptides.
 
@@
 
-The dataset contains files resulting from various stages of the benchmark:
+The dataset contains files resulting from various steps in the generation of the benchmark:
 \begin{itemize}
@@
 Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
+\fixme{Add Zenodo URLs.}
 
@@
   \centering
-  \begin{tabular}{ll}
-    \fixme{Insert figure} & \fixme{Insert figure} \\
-    a & b \\
-    \fixme{Insert figure} \\
-    c \\
+  \begin{tabular}{lll}
+    \includegraphics[height=1.4in]{nine-species-fdr.pdf} &
+    \includegraphics[height=1.4in]{percent-id.pdf}  &
+    \includegraphics[height=1.4in]{match-by.pdf}  \\
+    a & b & c \\
   \end{tabular}
   \caption{\textbf{Validation of the benchmarks.}
-    \textbf{a.} Each series indicates the number of accepted PSMs at 1\% PSM-level FDR for a given species.
+    \textbf{a.} Each series indicates, for a given species, the number of accepted PSMs as a function of PSM-level FDR.
     \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
-    \textbf{c.} Each violin plot indicates the distribution of the proportion of total ion current that is explained by PSMs per species.}
+    \textbf{c.} Each histogram show the distribution, per species, of the proportion of total ion current that is explained per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
   \label{fig:fdr}
@@
 To assess the overall rate of successful identification of spectra in each data set, we plotted the number of accepted PSMs as a function of PSM-level FDR threshold (Figure~\ref{fig:fdr}a).
-\fixme{Describe plot.}
-We also computed the proportion of spectra that were accepted at 1\% PSM-level FDR per species (Figure~\ref{fig:fdr}b).
-\fixme{Describe plot.}
-In addition, for each PSM we computed the proportion of the total peak intensities that is explained by matched b- and y-ions, using a matching tolerance of 0.05~\textit{m/z}.
-The resulting distributions of proportions differ substantially by species (Figure~\ref{fig:fdr}c).
-\fixme{Describe plot.}
+As is typical in proteomics database search, the curves go up rapidly before leaving the y-axis, corresponding to the many spectra with highly confident peptide assignments.
+To better understand the relative quality of the datasets, we also computed the proportion of spectra that were accepted at 1\% PSM-level FDR per species (Figure~\ref{fig:fdr}b).
+Here we observe that some datasets yield much higher rates of accepted PSMs than others, up to 39.7\% for \textit{Saccharomyces cerevisiae} and down to 3.6\% for \textit{Candidatus endoloripes}.
+Despite this large variance in the rate of accepted PSMs, characterizing the proportion of the total peak intensities that is explained by matched b- and y-ions (Figure~\ref{fig:fdr}c) suggests that the quality of the accepted PSMs is consistent across species.
 

### 2024-07-24 doc/nine-species/nine-species.tex
Edits

@@
     \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
-    \textbf{c.} Each histogram show the distribution, per species, of the proportion of total ion current that is explained per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
+    \textbf{c.} Each histogram show the distribution, per species, of the proportion of total ion current that is matched by b- or y-ions per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
   \label{fig:fdr}

### 2024-07-24 results/wnoble/2024-07-17validations/make_histograms.py

@@ -2,6 +2,7 @@
 # AUTHOR: WSN
 # CREATE DATE: 17 July 2024
 import sys
+import re
 import matplotlib.pyplot as plt
 
 ###########################################################################
@@ -27,10 +28,11 @@ def main():
 
     # Make the histogram.
     fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)
+    axs.xlabel("Percent matched")
     i = 0
     for species in match_df.keys():
         axs[i].hist(match_df[species], density=True, bins=100)
-        axs[i].set_title(species, loc='right')
+        axs[i].set_title(re.sub("-", " ", species), loc='right')
         axs[i].spines['top'].set_visible(False)
         axs[i].spines['right'].set_visible(False)
         i += 1

### 2024-07-24 results/wnoble/2024-07-17validations/make_histograms.py
Edits

@@ -17,7 +17,8 @@ def main():
     # Read all the match percentages from species-specific files.
     match_df = {} # Key = species, value = list of match percentages.
     for match_filename in sys.argv[2:]:
-        species = match_filename.split(".")[1] # match_by.<species>.txt
+        # match_by.<species>.txt (H.-sapiens special case)
+        species = ".".join(match_filename.split(".")[1:-1]) 
         with open(match_filename, "r") as match_file:
             match_file.readline()
             match_df[species] = [float(line.rstrip().split("\t")[2])
@@ -26,7 +27,8 @@ def main():
                   f"{species}.", file=sys.stderr)
 
     # Make the histogram.
-    fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)
+    fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True,
+                            figsize=(4,8))
     i = 0
     for species in match_df.keys():
         axs[i].hist(match_df[species], density=True, bins=100)

### 2024-07-24 results/wnoble/2024-07-17validations/make_fdr_curves.py
Add missing script

@@ -1,0 +1,45 @@
+#!/usr/bin/env python
+# AUTHOR: WSN
+# CREATE DATE: 17 July 2024
+import sys
+import pandas
+import re
+import os.path
+import matplotlib.pyplot as plt
+
+###########################################################################
+# MAIN
+###########################################################################
+def main():
+
+    # Check the command line.
+    if len(sys.argv) < 3:
+        print("USAGE: make_fdr_curves.py <plot> <percolator>+",
+              file=sys.stderr)
+        sys.exit(1)
+
+    fig = plt.figure(figsize=(4,8))
+    ax = plt.gca()
+    ax.set_xlim(0, 0.1)
+    ax.set_ylim(0, 2e6)
+    plt.xlabel("FDR threshold")
+    plt.ylabel("Accepted PSMs")
+
+    i = 0
+    for percolator_filename in sys.argv[2:]:
+        species = re.sub(
+            "-", " ", os.path.split(os.path.dirname(percolator_filename))[-1]
+        )
+        print(f"Reading {species}.", file=sys.stderr)
+
+        percolator_df = pandas.read_csv(percolator_filename, sep="\t")
+        qvalues = percolator_df["q-value"].values.tolist()
+
+        plt.plot(qvalues, range(len(qvalues)), label=species)
+        i += 1
+
+    plt.legend(loc='center right')
+    plt.savefig(sys.argv[1])
+
+if __name__ == "__main__":
+    main()

### 2024-07-24 doc/nine-species/nine-species.tex
Edits
Various minor updates

--- 2024-07-20 doc/nine-species/nine-species.tex
@@
   \begin{tabular}{lll}
-    \includegraphics[height=1.4in]{nine-species-fdr.pdf} &
-    \includegraphics[height=1.4in]{percent-id.pdf}  &
-    \includegraphics[height=1.4in]{match-by.pdf}  \\
+    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/nine-species-fdrs.pdf} &
+    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/percent-id.pdf}  &
+    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/match-by.pdf}  \\
     a & b & c \\
@@
     \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
-    \textbf{c.} Each histogram show the distribution, per species, of the proportion of total ion current that is explained per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
+    \textbf{c.} Each histogram shows, for one species, the distribution of the proportion of total ion current that is matched by b- or y-ions per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
   \label{fig:fdr}

### 2024-07-24 results/wnoble/2024-07-17validations/make_barchart.py

@@ -32,8 +32,10 @@ def main():
     print(percents, file=sys.stderr)
 
     # Make the barchart.
-    plt.barh(species, percents)
-    plt.xlabel("Percent identified")
+    fig, ax = plt.subplots()
+    fig.set_size_inches(4,8)
+    ax.barh(species, percents)
+    ax.set_xlabel("Percent identified")
     plt.tight_layout()
     plt.savefig(sys.argv[2])
 

### 2024-07-24 results/wnoble/2024-07-17validations/make_fdr_curves.py

@@ -18,7 +18,7 @@ def main():
               file=sys.stderr)
         sys.exit(1)
 
-    fig = plt.figure(figsize=(4,8))
+    fig = plt.figure(figsize=(4,8), tight_layout=True)
     ax = plt.gca()
     ax.set_xlim(0, 0.1)
     ax.set_ylim(0, 2e6)

### 2024-07-24 results/wnoble/2024-07-17validations/make_histograms.py

@@ -28,16 +28,17 @@ def main():
                   f"{species}.", file=sys.stderr)
 
     # Make the histogram.
-    fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True,
-                            subplot_kw={(figsize:(4,8))})
-    axs.xlabel("Percent matched")
+    fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)
+    fig.set_size_inches(4,8)
     i = 0
     for species in match_df.keys():
         axs[i].hist(match_df[species], density=True, bins=100)
         axs[i].set_title(re.sub("-", " ", species), loc='right')
+        axs[i].set_ylim(0,0.1)
         axs[i].spines['top'].set_visible(False)
         axs[i].spines['right'].set_visible(False)
         i += 1
+    axs[i-1].set_xlabel("Percent matched")
     plt.savefig(sys.argv[1])
 
 if __name__ == "__main__":

### 2024-07-25 results/wnoble/2024-07-17validations/make_histograms.py
Updates

@@ -34,6 +34,7 @@ def main():
     for species in match_df.keys():
         axs[i].hist(match_df[species], density=True, bins=100)
         axs[i].set_title(re.sub("-", " ", species), loc='right')
+        axs[i].set_xlim(0,60)
         axs[i].set_ylim(0,0.1)
         axs[i].spines['top'].set_visible(False)
         axs[i].spines['right'].set_visible(False)

### 2024-07-25 bin/downsample_benchmark.py
Fix downsampling script

@@ -40,8 +40,12 @@ def main():
     os.makedirs(args.root, exist_ok=True)
 
     for mgf_dir in args.mgf_dirs:
-        os.makedirs(os.path.join(args.root, os.path.basename(mgf_dir)),
-                    exist_ok=True)
+
+        out_dir = os.path.join(args.root, os.path.basename(mgf_dir))
+        # N.B. Hack to skip existing files.
+        if os.path.isfile(out_dir):
+            continue
+        os.makedirs(out_dir, exist_ok=True)
         mgf_filenames = glob.glob(os.path.join(mgf_dir, "*.mgf"))
         random.shuffle(mgf_filenames)
         total_spectra = 0

### 2024-07-26 bin/downsample_benchmark.py
Tiny fix

@@ -44,6 +44,7 @@ def main():
         out_dir = os.path.join(args.root, os.path.basename(mgf_dir))
         # N.B. Hack to skip existing files.
         if os.path.isfile(out_dir):
+            print(f"Skipping {mgf_dir}.", file=sys.stderr)
             continue
         os.makedirs(out_dir, exist_ok=True)
         mgf_filenames = glob.glob(os.path.join(mgf_dir, "*.mgf"))

### 2024-07-26 doc/nine-species/cover-letter.tex
Fix colors in plots

\documentclass[12pt]{letter}
\usepackage[margin=1in]{geometry}
\usepackage{url}
\usepackage[colorlinks]{hyperref}
\usepackage{graphicx}

\name{William Stafford Noble, Professor\\
Department of Genome Sciences\\
University of Washington}

\thispagestyle{empty}

\begin{document}

\begin{letter}{Dr.\ Guy Jones, Senior Editor \\
{\em Scientific Data}}

\opening{Dear Dr.\ Jones,}

I am pleased to submit a manuscript, ``A multi-species benchmark for training and validating large scale mass spectrometry proteomics machine learning models,'' for consideration for publication in {\em Scientific Data}.

This manuscript describes a dataset that was curated for development of our Casanovo {\em de novo} peptide sequencing model.
The model is described in the paper, ``Sequence-to-sequence translation from mass spectra to peptides with a transformer model,'' which is currently in press in {\em Nature Communications}.
The dataset consists of annotated peptide mass spectra derived from nine different species.
This nine-species benchmark is modeled after the dataset used in a seminal paper from 2017, ``{\em De novo} peptide sequencing by deep learning,'' published by Tran {\em et al.}\ in {\em PNAS}.
In revisiting that benchmark, we identified a series of significant problems with the original processing pipeline, including incorrect search settings and incomplete separation of train and test sets.
These problems are described in more detail in the accompanying manuscript.
Our revised benchmark fixes these problems and makes the data, as well as intermediate files, available via Zenodo.

Please note that, in addition to this benchmark dataset, our forthcoming Casanovo paper makes use of a separate dataset containing 30 million human spectra, derived from the MassIVE-KB database.
This dataset was initially developed for our paper, ``A learned embedding for efficient joint analysis of millions of mass spectra,'' by Bittremieux {\em et al.}, published in {\em Nature Methods} in 2022.
We are preparing separate manuscript describing this dataset, which we intend to submit to {\em Scientific Data} in the coming weeks.

Thank you for considering our manuscript for review.

\closing{Best regards,}
\end{letter}

\end{document}

### 2024-07-26 doc/nine-species/nine-species.tex
Fix colors in plots

@@
 Here we observe that some datasets yield much higher rates of accepted PSMs than others, up to 39.7\% for \textit{Saccharomyces cerevisiae} and down to 3.6\% for \textit{Candidatus endoloripes}.
-Despite this large variance in the rate of accepted PSMs, characterizing the proportion of the total peak intensities that is explained by matched b- and y-ions (Figure~\ref{fig:fdr}c) suggests that the quality of the accepted PSMs is consistent across species.
+Despite this large variance in the rate of accepted PSMs, characterizing the proportion of the total peak intensities that is explained by matched b- and y-ions (Figure~\ref{fig:fdr}c) suggests that the quality of the accepted PSMs is high.
+Notably, the proportion of matched  b- and y-ions does not appear to be strongly correlated with the rate of accepted PSMs per species.
 
 \section*{Code Availability}
-All code required to generate the various benchmarks and to produce the figures in this manuscript is available with an Apache license at \url{https://github.com/Noble-Lab/nine-species}, with a snapshot of the repository stored at \fixme{zenodo}.
+All code required to generate the various benchmarks and to produce the figures in this manuscript is available with an Apache license at \url{https://github.com/Noble-Lab/multi-species-benchmark}, with a snapshot of the repository stored at \url{https://zenodo.org/doi/10.5281/zenodo.12926326}.
 
@@
 
-\paragraph{Author Contributions}
+%\paragraph{Author Contributions}
 

### 2024-07-26 results/wnoble/2024-07-17validations/make_barchart.py

@@ -34,7 +34,9 @@ def main():
     # Make the barchart.
     fig, ax = plt.subplots()
     fig.set_size_inches(4,8)
-    ax.barh(species, percents)
+    colors = ['tab:blue', 'tab:orange', 'tab:green', 'tab:red', 'tab:purple',
+              'tab:brown', 'tab:pink', 'tab:gray', 'tab:olive']
+    ax.barh(species, percents, color=colors[::-1])
     ax.set_xlabel("Percent identified")
     plt.tight_layout()
     plt.savefig(sys.argv[2])

### 2024-07-26 results/wnoble/2024-07-17validations/make_fdr_curves.py

@@ -25,6 +25,10 @@ def main():
     plt.xlabel("FDR threshold")
     plt.ylabel("Accepted PSMs")
 
+
+    colors = ['tab:blue', 'tab:orange', 'tab:green', 'tab:red', 'tab:purple',
+              'tab:brown', 'tab:pink', 'tab:gray', 'tab:olive']
+
     i = 0
     for percolator_filename in sys.argv[2:]:
         species = re.sub(
@@ -35,7 +39,7 @@ def main():
         percolator_df = pandas.read_csv(percolator_filename, sep="\t")
         qvalues = percolator_df["q-value"].values.tolist()
 
-        plt.plot(qvalues, range(len(qvalues)), label=species)
+        plt.plot(qvalues, range(len(qvalues)), label=species, color=colors[i])
         i += 1
 
     plt.legend(loc='center right')

### 2024-07-26 results/wnoble/2024-07-17validations/make_histograms.py

@@ -27,12 +27,15 @@ def main():
             print(f"Read {len(match_df[species])} match percentages from " +
                   f"{species}.", file=sys.stderr)
 
+    colors = ['tab:blue', 'tab:orange', 'tab:green', 'tab:red', 'tab:purple',
+              'tab:brown', 'tab:pink', 'tab:gray', 'tab:olive']
+
     # Make the histogram.
     fig, axs = plt.subplots(len(match_df), 1, sharex=True, tight_layout=True)
     fig.set_size_inches(4,8)
     i = 0
     for species in match_df.keys():
-        axs[i].hist(match_df[species], density=True, bins=100)
+        axs[i].hist(match_df[species], density=True, bins=100, color=colors[i])
         axs[i].set_title(re.sub("-", " ", species), loc='right')
         axs[i].set_xlim(0,60)
         axs[i].set_ylim(0,0.1)

### 2024-08-04 doc/nine-species/nine-species.tex
A few small updates

@@
 
-\title{A multi-species benchmark for training and validating large scale mass spectrometry proteomics machine learning models}
+\title{A multi-species benchmark for training and validating mass spectrometry proteomics machine learning models}
 
@@
 
+\begin{sloppypar}
 We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
@@
 At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 179,256 distinct peptides.
+\end{sloppypar}
 
@@
 Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
-\fixme{Add Zenodo URLs.}
+The data is available at \url{https://doi.org/10.5281/zenodo.12819175}.
 
@@
 
-%\paragraph{Author Contributions}
+\paragraph{Author Contributions}
+WSN prepared the dataset, performed the validation experiments, and wrote the manuscript.
 

### 2024-08-04 results/wnoble/2024-07-17validations/make_barchart.py

@@ -38,6 +38,7 @@ def main():
               'tab:brown', 'tab:pink', 'tab:gray', 'tab:olive']
     ax.barh(species, percents, color=colors[::-1])
     ax.set_xlabel("Percent identified")
+    ax.set_yticklabels([])
     plt.tight_layout()
     plt.savefig(sys.argv[2])
 

### 2024-08-20 bin/summarize_benchmark.py
Add draft of reviewer response.  Modify summarization script to list distinct peptides

@@ -39,9 +39,10 @@ def clean_peptide (peptide):
     return i2l(re.sub(r"[0-9\.\[\]]+", "", peptide))
     
 
-def count_peptides(mgf_filenames):
+def count_peptides(mgf_filenames, output_filename):
     """Count number of distinct peptides in a set of MGF files.  Eliminates
-    modifications and converts isoleucines first."""
+    modifications and converts isoleucines first.  Prints a list of peptides
+    to the given file."""
 
     peptides = {} # Key = peptide, value = True
     for mgf_filename in mgf_filenames:
@@ -49,6 +50,9 @@ def count_peptides(mgf_filenames):
             for line in mgf_file:
                 if line.startswith("SEQ="):
                     peptides[clean_peptide(line.rstrip().split("=")[1])] = True
+
+    with open(output_filename, "w") as output_file:
+        print(list(peptides.keys()).join("\n"), file=output_file)
 
     return len(peptides)
                             
@@ -95,7 +99,9 @@ def main():
         output["#mgf"].append(len(benchmark_mgfs))
         output["#spectra"].append(count_spectra(data_mgfs))
         output["#PSMs"].append(count_spectra(benchmark_mgfs))
-        output["#peptides"].append(count_peptides(benchmark_mgfs))
+        output["#peptides"].append(count_peptides(benchmark_mgfs),
+                                   os.path.join(args.benchmark_dir, species,
+                                                "peptides.txt"))
 
     output["precursor"] = list(driver[2])
     output["fragment"] = list(driver[3])

### 2024-08-20 bin/clean-benchmark.py
Fix bug that handled + and in peptides incorrectly.

@@ -48,7 +48,7 @@ def i2l(peptide):
 def clean_peptide (peptide, do_i2l):
     "Remove flanking amino acids and PTMs."
 
-    no_mods = re.sub(r"[0-9\.\[\]]+", "", peptide)
+    no_mods = re.sub(r"[0-9\.\[\]\+\-]+", "", peptide)
 
     if do_i2l:
         return i2l(no_mods)

### 2024-08-20 bin/summarize_benchmark.py

@@ -36,7 +36,7 @@ def i2l(peptide):
 def clean_peptide (peptide):
     "Remove PTMs."
 
-    return i2l(re.sub(r"[0-9\.\[\]]+", "", peptide))
+    return i2l(re.sub(r"[0-9\.\[\]\+\-]+", "", peptide))
     
 
 def count_peptides(mgf_filenames, output_filename):
@@ -52,7 +52,8 @@ def count_peptides(mgf_filenames, output_filename):
                     peptides[clean_peptide(line.rstrip().split("=")[1])] = True
 
     with open(output_filename, "w") as output_file:
-        print(list(peptides.keys()).join("\n"), file=output_file)
+        print(f"Printing peptides to {output_filename}.\n", file=sys.stderr)
+        print("\n".join(sorted(list(peptides.keys()))), file=output_file)
 
     return len(peptides)
                             
@@ -99,9 +100,11 @@ def main():
         output["#mgf"].append(len(benchmark_mgfs))
         output["#spectra"].append(count_spectra(data_mgfs))
         output["#PSMs"].append(count_spectra(benchmark_mgfs))
-        output["#peptides"].append(count_peptides(benchmark_mgfs),
-                                   os.path.join(args.benchmark_dir, species,
-                                                "peptides.txt"))
+        output["#peptides"].append(
+            count_peptides(benchmark_mgfs,
+                           os.path.join(args.benchmark_dir, species,
+                                        "peptides.txt"))
+        )
 
     output["precursor"] = list(driver[2])
     output["fragment"] = list(driver[3])

### 2024-08-20 doc/nine-species/response.tex
Add reviewer response

\documentclass{article}
\usepackage[margin=1in]{geometry} % Set margins.
\usepackage{url} % Allow URLs.
\usepackage{graphicx} % Allow figures
\usepackage{amsmath} % Allow fancy math.
\usepackage{amssymb}
\usepackage{algorithm, algpseudocode} % Format algorithms
\usepackage[sort&compress]{natbib} % Better bibliography formatting.
\usepackage{xcolor} % Allow colored text.
\usepackage{siunitx}

% Turn off indenting.
\setlength{\parindent}{0pt}
\setlength\parskip{0.5em plus 0.1em minus 0.2em}

\newcommand{\response}[1]{\vspace*{1ex} \color{blue} \noindent #1 \color{black}
	\vspace*{2ex}}

\newcommand{\edit}[1]{\begin{quotation}\color{red}\noindent #1
		\color{black}\end{quotation}}

\newcommand{\myinput}[1]{\begin{quotation}\noindent\color{red}\input{#1}\color{black}\end{quotation}
\vspace*{1ex}
}

\usepackage{soul}
\newcommand{\fixme}[1]{\hl{#1}}

\begin{document}

\noindent
\today\\[2ex]

\noindent
Dear Dr.\ Guo:\\[2ex]

We thank you and the reviewers for your kind consideration of our manuscript.
Below, we address each of the points raised by the two reviewers and describe the changes we have made to the manuscript.
In what follows, the reviewer's comments are shown in black type, interleaved with our responses (in blue) and, where appropriate, the modified text (in red).
We have also made the formatting changes that you requested.

Thank you very much for your consideration.\\[2ex]

\noindent
Best regards,\\[2ex]

\noindent
William Noble\\
University of Washington

\clearpage
\section*{Reviewer 1}

The author pointed out a significant issue in de novo peptide sequencing. I really appreciate his efforts to provide this curated benchmark dataset for the proteomics community. This kind of benchmark dataset is valuable and useful for training high-quality models. However, I have several technical concerns. Here they are:

(1) The author reanalyzed the nine-species dataset using Tide search engine. I highly suggest at least two search engines should be used to make sure the PSMs are correct, since almost every search engine has its own preference.

\response{We agree with you that the choice of search engine introduces a potential bias into the dataset construction.
  However, mitigating this bias is challenging.
  If we combine the results of two different search engine, using either a union or an intersection operation, then we sacrifice the ability to rigorously control the FDR among the reported list of discoveries.}

(2) Why did the author only consider the FDR at PSM level? I think the FDR at protein level is commonly used now.

\response{We agree that, in many settings, controlling the FDR at the protein level is appropriate, especially when a study is aimed at characterizing the protein content of a given set of biological samples.
  However, for constructing a benchmark of PSMs used for training and validating \textit{de novo} sequencing methods, we believe that protein level FDR control is not as relevant.
  In particular, even if we control the FDR at the protein level, the benchmark needs to report a set of PSMs rather than proteins, and it is not clear how to go from a set of detected proteins to a corresponding collection of confident PSMs.
  Consider, for example, the widely used ``picked protein'' method for protein-level FDR control \cite{savitski2015scalable}.
  In the Savitski \textit{et al.}\ (2015) description of this method, each protein is assigned a score equal to the sum of the search engine scores assigned to its peptides.
  With such an approach, a protein can be detected with high confidence even if one or more of its associated peptides has low scores, as long as some of the peptides have very high scores.
  Thus, if we include all PSMs associated with a detected peptide in our benchmark, we risk contaminating the benchmark with incorrect PSMs.}

(3) The annotation in MGF is adding the peptide sequence. Is it possible to provide the annotated product ions? For example, which peak is annotated as b5 ion or y3 ion. With these annotations at ion level, this dataset will be more useful.

\response{Yes, we have added to the benchmark a \fixme{what format?} file that provides this annotation.  Here is the corresponding description from the revised manuscript:}

\fixme{[Bill] Write this.}

\clearpage
\section*{Reviewer 2}

This manuscript describes a new dataset designed for evaluating de novo peptide sequencing algorithms. This new dataset represents an enhancement of the widely-used nine-species dataset, addressing two key issues present in the original dataset:

1. The nine-species dataset incorrectly assigns deamidation modifications to isotopic peaks too frequently.

2. In the nine-species dataset, some peptides are shared across species, which can lead to overly optimistic evaluation outcomes for complex models.

Additionally, a balanced dataset is included for specialized evaluation purposes.

Overall, the dataset generation process is clearly outlined, but some important statistical numbers reported about the generation process need clarification. Specifically, on page 4, in the section titled ``Database Search and FDR Control," the manuscript states that ``At this point... the dataset contains 179,256 distinct peptides." Then, in the next section, it mentions ``Among the 229,984 unique peptides," and after further processing, ``The final, non-redundant benchmark dataset (`main') consists of... 258,105 distinct peptides." It is unclear how these three numbers of peptides are related, and what distinguishes ``unique peptides" from ``distinct peptides." Clarifying these relationships and definitions is essential for assessing the quality of the dataset generation process.

\response{We apologize for the confusion, and we thank the reviewer for pointing out the inconsistency in the reported numbers of peptides in the benchmark.
  We have fixed a bug in the way that our scripts handle PTM annotations in sequences, and the tables in the manuscript as well as the Zenodo files have been updated accordingly.
  \fixme{[Bill] Finish making these changes.}
  We previously used the term ``unique'' and ``distinct'' interchangeably, and in the revised version of the manuscript we stick to ``distinct.''
}

\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}

### 2024-08-21 doc/nine-species/response.tex
Update benchmark numbers

@@
   We have fixed a bug in the way that our scripts handle PTM annotations in sequences, and the tables in the manuscript as well as the Zenodo files have been updated accordingly.
-  \fixme{[Bill] Finish making these changes.}
+  In the new version, there are 168,422 distinct peptides after the initial construction of the benchmark.
+  This number remains the same after eliminating overlaps (since each distinct peptide is randomly assigned to a single species, as described in the text).
+  The downsampled benchmark contains 133,232 distinct peptides.
+
   We previously used the term ``unique'' and ``distinct'' interchangeably, and in the revised version of the manuscript we stick to ``distinct.''

### 2024-08-21 doc/nine-species/nine-species.tex
Add draft of reviewer response.  Modify summarization script to list distinct peptides
Update benchmark numbers

@@
 
-\author[1,2]{William Stafford Noble}
+\author[1]{Wout Bittremieux}
+\author[2,3]{William Stafford Noble\footnote{Correspondence: william-noble@uw.edu}}
 
-\affil[1]{Department of Genome Sciences, University of Washington}
-\affil[2]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington}
+\affil[1]{Department of Computer Science, University of Antwerp, Belgium}
+\affil[2]{Department of Genome Sciences, University of Washington, Seattle, WA, USA}
+\affil[3]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington, Seattle, WA, USA}
 
@@
 \maketitle
-
-%Instructions at
-%  \url{https://www.nature.com/sdata/publish/submission-guidelines}.}
 
@@
 PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
-108,756 & 13,768 & 102,576 & 13,656 & 20 & 0.05 \\
+108,402 & 11,638 & 102,255 & 11,557
+& 20 & 0.05 \\
 PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
-25,490 & 6833 & 25,490 & 6833 & 10 & 0.05 \\
+25,522 & 5630 & 25,522 & 5630
+& 10 & 0.05 \\
 PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
-267,269 & 23,448 & 100,517 & 16,774 & 10 & 0.05 \\
+267,183 & 15,220 & 100,485 & 11,934
+& 10 & 0.05 \\
 PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
-1,357,908 & 64,025 & 113,613 & 29,245 & 30 & 0.05 \\
+1,351,938 & 28,364 & 113,234 & 17,481
+& 30 & 0.05 \\
 PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
-82,204 & 10,863 & 82,204 & 10,863 & 20 & 0.05 \\
+82,514 & 8080 & 82,514 & 8080
+& 20 & 0.05 \\
 PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
-177,941 & 62,294 & 100,325 & 44,052 & 15 & 0.05 \\
+177,553 & 48,459 & 100,056 & 35,787
+& 15 & 0.05 \\
 PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
-586,278 & 33,008 & 109,023 & 20,057 & 20 & 0.05 \\
+585,846 & 19,102 & 108,973 & 13,285
+& 20 & 0.05 \\
 PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
-193,437 & 30,013 & 101,700 & 24,905 & 20 & 0.05 \\
+194,604 & 21,081 & 102,285 & 18,630
+& 20 & 0.05 \\
 PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
-44,348 & 13,853 & 44,348 & 13,853 & 20 & 0.02 \\
+44,555 & 10,848 & 44,555 & 10,848
+& 20 & 0.02 \\
 \hline
-Total & & & 343 & 15,165,161  & 2,843,631 & 258,105 & 779,796 & 180,238  \\
+Total & & & 343 & 15,165,161  & 2,838,117 & 168,422 & 779,879 & 133,232 \\
 \end{tabular}
@@
 We identified 13 MGF files with fewer than 100 accepted PSMs, and we eliminated all of these PSMs from the benchmark.
-At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 179,256 distinct peptides.
+At this point in the processing pipeline, the dataset contains 2,898,611 annotated spectra (PSMs)  drawn from 339 RAW files and associated with 168,422 distinct peptides.
 \end{sloppypar}
@@
 To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
-Among the 229,984 unique peptides, we identified 3797 (1.7\%) that occur in more than one species.
+Among the 168,422 distinct peptides, we identified 4121 (2.4\%) that occur in more than one species.
 For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
@@
 Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
-The final, non-redundant benchmark dataset (``main'') consists of 2,843,631 PSMs corresponding to 258,105 distinct peptides.
+The final, non-redundant benchmark dataset (``main'') consists of 2,838,117 PSMs corresponding to 168,422 distinct peptides.
 
@@
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,796 PSMs from 180,238 distinct peptides.
+The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,879 PSMs from 133,232 distinct peptides.
 

### 2024-08-27 bin/summarize_benchmark.py
Update figs

@@ -52,7 +52,6 @@ def count_peptides(mgf_filenames, output_filename):
                     peptides[clean_peptide(line.rstrip().split("=")[1])] = True
 
     with open(output_filename, "w") as output_file:
-        print(f"Printing peptides to {output_filename}.\n", file=sys.stderr)
         print("\n".join(sorted(list(peptides.keys()))), file=output_file)
 
     return len(peptides)

### 2024-08-29 doc/nine-species/response.tex
Update response

@@
 
-\response{Yes, we have added to the benchmark a \fixme{what format?} file that provides this annotation.  Here is the corresponding description from the revised manuscript:}
-
-\fixme{[Bill] Write this.}
+\response{Yes, we have added to the benchmark, alongside each MGF file a corresponding mzSpecLib \cite{klein2024proteomics} file that provides this annotation.}
 

### 2024-08-29 doc/nine-species/nine-species.tex
Update response

@@
 \author[1]{Wout Bittremieux}
+\author[2]{Bo Wen}
 \author[2,3]{William Stafford Noble\footnote{Correspondence: william-noble@uw.edu}}
@@
 \item PSM-level Percolator results files for targets, in tab-delimited format.
-\item Annotated MGF files for both versions of the benchmark (main and balanced).
+`\item Annotated MGF and corresponding mzSpecLib \cite{klein2024proteomics} files for both versions of the benchmark (main and balanced).
 \end{itemize}
@@
   \begin{tabular}{lll}
-    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/nine-species-fdrs.pdf} &
-    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/percent-id.pdf}  &
-    \includegraphics[height=4in]{../../results/wnoble/2024-07-17validations/match-by.pdf}  \\
+    \includegraphics[height=4in]{nine-species-fdrs.pdf} &
+    \includegraphics[height=4in]{percent-id.pdf}  &
+    \includegraphics[height=4in]{match-by.pdf}  \\
     a & b & c \\

### 2024-08-29 doc/nine-species/response.tex
Edit

@@
 Below, we address each of the points raised by the two reviewers and describe the changes we have made to the manuscript.
-In what follows, the reviewer's comments are shown in black type, interleaved with our responses (in blue) and, where appropriate, the modified text (in red).
-We have also made the formatting changes that you requested.
+In what follows, the reviewer's comments are shown in black type, interleaved with our responses in blue.
 

### 2024-08-29 doc/nine-species/nine-species.tex
Edit

@@
 
-\newcommand{\fixme}[1]{{\color{red}{\bf #1}\color{black}}}
+\newcommand{\edit}[1]{{\color{red}{#1}\color{black}}}
 
@@
 
-\author[1]{Wout Bittremieux}
-\author[2]{Bo Wen}
-\author[2,3]{William Stafford Noble\footnote{Correspondence: william-noble@uw.edu}}
+\author[1]{Bo Wen}
+\author[1,2]{William Stafford Noble}
 
-\affil[1]{Department of Computer Science, University of Antwerp, Belgium}
-\affil[2]{Department of Genome Sciences, University of Washington, Seattle, WA, USA}
-\affil[3]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington, Seattle, WA, USA}
+\affil[1]{Department of Genome Sciences, University of Washington, Seattle, WA, USA}
+\affil[2]{Paul G.\ Allen School of Computer Science and Engineering, University of Washington, Seattle, WA, USA}
 
@@
 \maketitle
+
+\noindent
+Correspondence: william-noble@uw.edu
 
@@
 PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
-108,402 & 11,638 & 102,255 & 11,557
+\edit{108,402} & \edit{11,638} & \edit{102,255} & \edit{11,557}
 & 20 & 0.05 \\
 PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
-25,522 & 5630 & 25,522 & 5630
+\edit{25,522} & \edit{5630} & \edit{25,522} & \edit{5630}
 & 10 & 0.05 \\
 PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
-267,183 & 15,220 & 100,485 & 11,934
+\edit{267,183} & \edit{15,220} & \edit{100,485} & \edit{11,934}
 & 10 & 0.05 \\
 PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
-1,351,938 & 28,364 & 113,234 & 17,481
+\edit{1,351,938} & \edit{28,364} & \edit{113,234} & \edit{17,481}
 & 30 & 0.05 \\
 PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
-82,514 & 8080 & 82,514 & 8080
+\edit{82,514} & \edit{8080} & \edit{82,514} & \edit{8080}
 & 20 & 0.05 \\
 PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
-177,553 & 48,459 & 100,056 & 35,787
+\edit{177,553} & \edit{48,459} & \edit{100,056} & \edit{35,787}
 & 15 & 0.05 \\
 PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
-585,846 & 19,102 & 108,973 & 13,285
+\edit{585,846} & \edit{19,102} & \edit{108,973} & \edit{13,285}
 & 20 & 0.05 \\
 PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
-194,604 & 21,081 & 102,285 & 18,630
+\edit{194,604} & \edit{21,081} & \edit{102,285} & \edit{18,630}
 & 20 & 0.05 \\
 PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
-44,555 & 10,848 & 44,555 & 10,848
+\edit{44,555} & \edit{10,848} & \edit{44,555} & \edit{10,848}
 & 20 & 0.02 \\
 \hline
-Total & & & 343 & 15,165,161  & 2,838,117 & 168,422 & 779,879 & 133,232 \\
+Total & & & 343 & 15,165,161  & \edit{2,838,117} & \edit{168,422} & \edit{779,879} & \edit{133,232} \\
 \end{tabular}
@@
 To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
-Among the 168,422 distinct peptides, we identified 4121 (2.4\%) that occur in more than one species.
+Among the \edit{168,422 distinct} peptides, we identified \edit{4121 (2.4\%)} that occur in more than one species.
 For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
@@
 Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
-The final, non-redundant benchmark dataset (``main'') consists of 2,838,117 PSMs corresponding to 168,422 distinct peptides.
+The final, non-redundant benchmark dataset (``main'') consists of \edit{2,838,117} PSMs corresponding to \edit{168,422} distinct peptides.
 
@@
 We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
-Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,541.
+Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with \edit{25,522}.
 Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
 To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
-This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,604 for \textit{H.\ sapiens} and 82,290 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
+This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (\edit{44,555} for \textit{H.\ sapiens} and \edit{82,514} for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,879 PSMs from 133,232 distinct peptides.
+The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of \edit{779,879} PSMs from \edit{133,232} distinct peptides.
 
@@
 \item PSM-level Percolator results files for targets, in tab-delimited format.
-`\item Annotated MGF and corresponding mzSpecLib \cite{klein2024proteomics} files for both versions of the benchmark (main and balanced).
+`\item Annotated MGF \edit{and corresponding mzSpecLib \cite{klein2024proteomics}} files for both versions of the benchmark (main and balanced).
 \end{itemize}
@@
 WSN prepared the dataset, performed the validation experiments, and wrote the manuscript.
+\edit{BW edited the manuscript and wrote the code to convert MGF to mzSpecLib.}
 

### 2024-08-30 bin/mgf_annotate/src/main/java/util/JSpectrum.java
Uploaded java code for spectral library generation.

@@ -1,0 +1,15 @@
+package main.java.util;
+
+import com.compomics.util.experiment.mass_spectrometry.spectra.Precursor;
+import com.compomics.util.experiment.mass_spectrometry.spectra.Spectrum;
+
+public class JSpectrum extends Spectrum {
+
+    public String title;
+    public String SEQ;
+    public int SCANS;
+
+    public JSpectrum(Precursor precursor, double[] mzArray, double[] intensityArray) {
+        super(precursor, mzArray, intensityArray);
+    }
+}

### 2024-08-30 bin/mgf_annotate/src/main/java/util/MGFUtils.java

@@ -1,0 +1,461 @@
+package main.java.util;
+
+
+import com.compomics.util.experiment.biology.ions.Ion;
+import com.compomics.util.experiment.biology.ions.NeutralLoss;
+import com.compomics.util.experiment.biology.ions.impl.PeptideFragmentIon;
+import com.compomics.util.experiment.biology.modifications.Modification;
+import com.compomics.util.experiment.biology.modifications.ModificationCategory;
+import com.compomics.util.experiment.biology.modifications.ModificationFactory;
+import com.compomics.util.experiment.biology.modifications.ModificationType;
+import com.compomics.util.experiment.biology.proteins.Peptide;
+import com.compomics.util.experiment.identification.matches.IonMatch;
+import com.compomics.util.experiment.identification.matches.ModificationMatch;
+import com.compomics.util.experiment.identification.spectrum_annotation.AnnotationParameters;
+import com.compomics.util.experiment.identification.spectrum_annotation.SpecificAnnotationParameters;
+import com.compomics.util.experiment.identification.spectrum_annotation.SpectrumAnnotator;
+import com.compomics.util.experiment.identification.spectrum_annotation.spectrum_annotators.PeptideSpectrumAnnotator;
+import com.compomics.util.experiment.identification.spectrum_assumptions.PeptideAssumption;
+import com.compomics.util.experiment.mass_spectrometry.spectra.Spectrum;
+import com.compomics.util.gui.waiting.waitinghandlers.WaitingHandlerCLIImpl;
+import com.compomics.util.parameters.identification.advanced.SequenceMatchingParameters;
+import com.compomics.util.parameters.identification.search.ModificationParameters;
+import com.compomics.util.waiting.WaitingHandler;
+import org.apache.commons.lang3.StringUtils;
+
+import java.io.BufferedWriter;
+import java.io.File;
+import java.io.FileWriter;
+import java.io.IOException;
+import java.util.ArrayList;
+import java.util.HashMap;
+import java.util.HashSet;
+import java.util.List;
+import java.util.regex.Matcher;
+import java.util.regex.Pattern;
+
+public class MGFUtils {
+
+    public double itol = 0.05;
+    public String itolu = "Da";
+    public boolean lossWaterNH3 = true;
+    public double fragment_ion_intensity_cutoff = 0.01;
+    public String out_dir = "./";
+    public boolean export_input_for_pdv = true;
+
[…]

### 2024-08-30 doc/nine-species/nine-species.tex
A few minor edits

@@
 \usepackage{url} % Allow URLs.
-\usepackage[sort&compress]{natbib} % Better bibliography formatting.
 
@@
 \hline
-PRIDE & Species & Uniprot & Files & Spectra &
+PRIDE & Species & UniProt & Files & Spectra &
 \multicolumn{2}{c}{Main} & \multicolumn{2}{c}{Balanced} & Pre & Frag \\
@@
 \item Paiva \textit{et al.}\ investigated the protein expression response of the cowpea plant (\textit{Vigna unguiculata}) to infection by \textit{Cowpsea severe mosaic virus} (CSMV) by carrying out label-free proteomic analysis of cowpea leaves that were inoculated with CSMV compared to mock inoculation controls \cite{paiva2016label}.
-\item Nevo \textit{et al.}\ studied a rare autosomal recessive lysosomal storage disorder, cystinosis, by carrying out SILAC proteomic analysis of engineered mouse cell lines that harbor a known pathenogenic mutation of the causative gene, \textit{CTNS} \cite{nevo2017impact}.
+\item Nevo \textit{et al.}\ studied a rare autosomal recessive lysosomal storage disorder, cystinosis, by carrying out SILAC proteomic analysis of engineered mouse cell lines that harbor a known pathogenic mutation of the causative gene, \textit{CTNS} \cite{nevo2017impact}.
 \item Cassidy \textit{et al.}\ evaluated two different analytical approaches for carrying out full proteome analysis while identifying short open reading frames: a high/low pH reversed phase LC-MS bottom-up approach and a semi-top-down strategy involving separation of proteins in a GelFree system followed by digestion and LC-MS analysis \cite{cassidy2016combination}.
@@
 We downloaded the RAW files from the corresponding PRIDE projects (Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4.
-We downloaded the corresponding nine Uniprot reference proteomes and constructed a Tide index for each one, using Crux version 4.2.
+We downloaded the corresponding nine UniProt reference proteomes and constructed a Tide index for each one, using Crux version 4.2.
 Note that, for one species (\textit{Vigna mungo}) no reference proteome is available, so we used the proteome of the closely related species \textit{Vigna radiata}.
@@
 We assigned peptide labels to spectra using the Tide search engine followed by post-processing with Percolator.
-In creating the Tide index, we specified specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
+In creating the Tide index, we specified Cys carbamidomethylation as a static modification and allowed for the following variable modifications: Met oxidation, Asn deamidation, Gln deamidation, N-term acetylation, N-term carbamylation, N-term NH$_{3}$ loss, and the combination of N-term carbamylation and NH$_{3}$ loss by using the tide-index options {\tt --mods-spec 1M+15.994915,\allowbreak 1N+0.984016,\allowbreak 1Q+0.984016 --nterm-peptide-mods-spec 1X+42.010565,\allowbreak 1X+43.005814,\allowbreak 1X-17.026549,\allowbreak 1X+25.980265 --max-mods 3}.
 Note that one of the nine experiments (\textit{Mus musculus}) was performed using SILAC labeling, but we searched without SILAC modifications and hence include in the benchmark only PSMs from unlabeled peptides.
@@
 \begin{itemize}
-\item Spectrum files in MGF format, produced by msconvert.
-\item Reference proteome files in FASTA format, downloaded from Uniprot.
+\item Spectrum files in MGF format, produced by ThermoRawFileParser.
+\item Reference proteome files in FASTA format, downloaded from UniProt.
 \item Search results files for both targets and decoys, in tab-delimited format, produced by Tide.
@@
 
-\bibliographystyle{unsrt}
+\bibliographystyle{naturemag}
 \bibliography{refs} % See https://github.com/Noble-Lab/noble-lab-references

### 2024-08-30 doc/nine-species/response.tex
Add a bit more to the response

@@
   With such an approach, a protein can be detected with high confidence even if one or more of its associated peptides has low scores, as long as some of the peptides have very high scores.
-  Thus, if we include all PSMs associated with a detected peptide in our benchmark, we risk contaminating the benchmark with incorrect PSMs.}
+  Thus, if we include all PSMs associated with a detected peptide in our benchmark, we risk contaminating the benchmark with incorrect PSMs.
+
+  It is possible that the reviewer is suggesting that we should control the FDR at both the PSM and protein level.
+  This approach would have the converse problem: some PSMs that are correctly detected might be discarded if there is insufficient evidence to accept the protein at the specified FDR threshold.
+  Ultimately, we are creating a benchmark of PSMs, so PSM-level FDR control seemed the most appropriate.
+}
 

### 2024-09-30 doc/nine-species/nine-species.tex
Added data citation

@@
 
-The dataset contains files resulting from various steps in the generation of the benchmark:
+The dataset \cite{noble2024benchmark} contains files resulting from various steps in the generation of the benchmark:
 \begin{itemize}
@@
 Also included are log files for the steps of the analysis pipeline carried out using Crux \cite{park2008rapid} (Tide indexing, Tide search, and Percolator).
-The data is available at \url{https://doi.org/10.5281/zenodo.12819175}.
+The data \cite{noble2024benchmark} is available at \url{https://doi.org/10.5281/zenodo.13685813}.
 

### 2024-09-30 doc/nine-species/nine-species.tex
Added citations to table 1

@@
 \hline
-PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
+PXD005025 \cite{paiva2016label} & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
 \edit{108,402} & \edit{11,638} & \edit{102,255} & \edit{11,557}
 & 20 & 0.05 \\
-PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
+PXD004948 \cite{nevo2017impact} & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
 \edit{25,522} & \edit{5630} & \edit{25,522} & \edit{5630}
 & 10 & 0.05 \\
-PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
+PXD004325 \cite{cassidy2016combination} & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
 \edit{267,183} & \edit{15,220} & \edit{100,485} & \edit{11,934}
 & 10 & 0.05 \\
-PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
+PXD004565 \cite{reuss2017large} & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
 \edit{1,351,938} & \edit{28,364} & \edit{113,234} & \edit{17,481}
 & 30 & 0.05 \\
-PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
+PXD004536 \cite{petersen2016chemosynthetic} & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
 \edit{82,514} & \edit{8080} & \edit{82,514} & \edit{8080}
 & 20 & 0.05 \\
-PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
+PXD004947 \cite{mata2017depth} & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
 \edit{177,553} & \edit{48,459} & \edit{100,056} & \edit{35,787}
 & 15 & 0.05 \\
-PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
+PXD003868 \cite{seidel2017quantitative} & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
 \edit{585,846} & \edit{19,102} & \edit{108,973} & \edit{13,285}
 & 20 & 0.05 \\
-PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
+PXD004467 \cite{hu2016proteome} & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
 \edit{194,604} & \edit{21,081} & \edit{102,285} & \edit{18,630}
 & 20 & 0.05 \\
-PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
+PXD004424 \cite{cypryk2017proteomic} & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
 \edit{44,555} & \edit{10,848} & \edit{44,555} & \edit{10,848}
@@
 
-We downloaded the RAW files from the corresponding PRIDE projects (Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4.
+We downloaded the RAW files from the corresponding PRIDE projects (Table~\ref{tab:benchmark}) and converted them to MGF format using the ThermoRawFileParser v1.3.4 \cite{hulstaert:thermorawfileparser}.
 We downloaded the corresponding nine UniProt reference proteomes and constructed a Tide index for each one, using Crux version 4.2.

### 2024-10-01 doc/nine-species/nine-species.tex
Move floats to end.  Remove edit macro

--- 2024-08-30 doc/nine-species/nine-species.tex
@@
 \usepackage{url} % Allow URLs.
-
-\newcommand{\edit}[1]{{\color{red}{#1}\color{black}}}
 
@@
 
-%% Species & #raw & #mgf & #spectra & #PSMs & #peptides & #PSMs & #peptides & precursor & fragment
-%% \hline
-%% \textit{Vigna mungo} & 24 & 23 & 932848 & 108756 & 13768 & 102576 & 13656 & 20 & 0.05
-%% \textit{Mus musculus} & 13 & 9 & 306786 & 25490 & 6833 & 25490 & 6833 & 10 & 0.05
-%% \textit{Methanosarcina mazei} & 72 & 26 & 3728183 & 267269 & 23448 & 100517 & 16774 & 10 & 0.05
-%% \textit{Bacillus subtilis} & 106 & 9 & 4336428 & 1357908 & 64025 & 113613 & 29245 & 30 & 0.05
-%% \textit{Candidatus endoloripes} & 11 & 11 & 2272023 & 82204 & 10863 & 82204 & 10863 & 20 & 0.05
-%% \textit{Solanum lycopersicum} & 60 & 33 & 603506 & 177941 & 62294 & 100325 & 44052 & 15 & 0.05
-%% \textit{Saccharomyces cerevisiae} & 27 & 5 & 1477397 & 586278 & 33008 & 109023 & 20057 & 20 & 0.05
-%% \textit{Apis mellifera} & 17 & 8 & 823169 & 193437 & 30013 & 101700 & 24905 & 20 & 0.05
-%% \textit{H. sapiens} & 26 & 26 & 684821 & 44348 & 13853 & 44348 & 13853 & 20 & 0.02
-%% Total & 356 & 150 & 15165161 & 2843631 & 258105 & 779796 & 180238 
-%% \hline
-
-%% \hline
-%% Species & #raw & #mgf & #spectra & #PSMs & #peptides & #PSMs & #peptides & precursor & fragment
-%% \hline
-%% \textit{Vigna mungo} & 24 & 23 & 932848 
-%% \textit{Mus musculus} & 13 & 9 & 306786 
-%% \textit{Methanosarcina mazei} & 72 & 26 &
-%% \textit{Bacillus subtilis} & 106 & 9 &
-%% \textit{Candidatus endoloripes} & 11 & 11 & 2272023 & 
-%% \textit{Solanum lycopersicum} & 60 & 33 & 603506 & 
-%% \textit{Saccharomyces cerevisiae} & 27 & 5 & 1477397 & 
-%% \textit{Apis mellifera} & 17 & 8 & 823169 & 
-%% \textit{H. sapiens} & 26 & 26 & 684821 & 
-%% Total & 356 & 150 & 15165161 & 
-%% \hline
-
 \begin{table}
@@
 PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
-\edit{108,402} & \edit{11,638} & \edit{102,255} & \edit{11,557}
+108,402 & 11,638 & 102,255 & 11,557
 & 20 & 0.05 \\
 PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
-\edit{25,522} & \edit{5630} & \edit{25,522} & \edit{5630}
+25,522 & 5630 & 25,522 & 5630
 & 10 & 0.05 \\
 PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
-\edit{267,183} & \edit{15,220} & \edit{100,485} & \edit{11,934}
+267,183 & 15,220 & 100,485 & 11,934
 & 10 & 0.05 \\
 PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
-\edit{1,351,938} & \edit{28,364} & \edit{113,234} & \edit{17,481}
+1,351,938 & 28,364 & 113,234 & 17,481
 & 30 & 0.05 \\
 PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
-\edit{82,514} & \edit{8080} & \edit{82,514} & \edit{8080}
+82,514 & 8080 & 82,514 & 8080
 & 20 & 0.05 \\
 PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
-\edit{177,553} & \edit{48,459} & \edit{100,056} & \edit{35,787}
+177,553 & 48,459 & 100,056 & 35,787
 & 15 & 0.05 \\
 PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
-\edit{585,846} & \edit{19,102} & \edit{108,973} & \edit{13,285}
+585,846 & 19,102 & 108,973 & 13,285
 & 20 & 0.05 \\
 PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
-\edit{194,604} & \edit{21,081} & \edit{102,285} & \edit{18,630}
+194,604 & 21,081 & 102,285 & 18,630
 & 20 & 0.05 \\
 PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
-\edit{44,555} & \edit{10,848} & \edit{44,555} & \edit{10,848}
+44,555 & 10,848 & 44,555 & 10,848
 & 20 & 0.02 \\
 \hline
-Total & & & 343 & 15,165,161  & \edit{2,838,117} & \edit{168,422} & \edit{779,879} & \edit{133,232} \\
+Total & & & 343 & 15,165,161  & 2,838,117 & 168,422 & 779,879 & 133,232 \\
 \end{tabular}
@@
 To avoid train/test leakage, we post-processed the PSMs to eliminate peptides that are shared between species.
-Among the \edit{168,422 distinct} peptides, we identified \edit{4121 (2.4\%)} that occur in more than one species.
+Among the 168,422 distinct peptides, we identified 4121 (2.4\%) that occur in more than one species.
 For each such peptide, we selected one of the associated species at random and then eliminated all PSMs containing that peptide in other species.
@@
 Hence, if a given peptide appears in more than one species, then that peptide, including all its modified forms, is randomly assigned to a single species and eliminated from the others.
-The final, non-redundant benchmark dataset (``main'') consists of \edit{2,838,117} PSMs corresponding to \edit{168,422} distinct peptides.
+The final, non-redundant benchmark dataset (``main'') consists of 2,838,117 PSMs corresponding to 168,422 distinct peptides.
 
@@
 We therefore used a random downsampling procedure to produce a benchmark that is more evenly balanced across species.
-Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with \edit{25,522}.
+Among the nine species, the one with the fewest PSMs is \textit{Mus musculus}, with 25,522.
 Downsampling all of the other eight species to have 25,000 PSMs would reduce the size of the dataset from 2.8 million PSMs to 225,000---a reduction of 92\%.
 To avoid producing such a small dataset, we therefore opted to downsample each dataset to approximately 100,000 PSMs.
-This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (\edit{44,555} for \textit{H.\ sapiens} and \edit{82,514} for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
+This approach yields a slight imbalance, because three species have fewer than 100,000 PSMs (44,555 for \textit{H.\ sapiens} and 82,514 for \textit{Candidatus endoloripes}), while retaining a larger percentage of the original data.
 Our downsampling procedure involved randomly permuting the order of the MGF files for each species and then selecting the files in order until at least 100,000 PSMs have been accepted.
-The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of \edit{779,879} PSMs from \edit{133,232} distinct peptides.
+The final, balanced benchmark dataset is approximately one quarter the size of the main benchmark, consisting of 779,879 PSMs from 133,232 distinct peptides.
 
@@
 \item PSM-level Percolator results files for targets, in tab-delimited format.
-`\item Annotated MGF \edit{and corresponding mzSpecLib \cite{klein2024proteomics}} files for both versions of the benchmark (main and balanced).
+`\item Annotated MGF and corresponding mzSpecLib \cite{klein2024proteomics} files for both versions of the benchmark (main and balanced).
 \end{itemize}
@@
 \section*{Technical Validation}
-
-\begin{figure}
-  \centering
-  \begin{tabular}{lll}
-    \includegraphics[height=4in]{nine-species-fdrs.pdf} &
-    \includegraphics[height=4in]{percent-id.pdf}  &
-    \includegraphics[height=4in]{match-by.pdf}  \\
-    a & b & c \\
-  \end{tabular}
-  \caption{\textbf{Validation of the benchmarks.}
-    \textbf{a.} Each series indicates, for a given species, the number of accepted PSMs as a function of PSM-level FDR.
-    \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
-    \textbf{c.} Each histogram shows, for one species, the distribution of the proportion of total ion current that is matched by b- or y-ions per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
-  \label{fig:fdr}
-\end{figure}
 
@@
 WSN prepared the dataset, performed the validation experiments, and wrote the manuscript.
-\edit{BW edited the manuscript and wrote the code to convert MGF to mzSpecLib.}
+BW edited the manuscript and wrote the code to convert MGF to mzSpecLib.
 
@@
 
+\paragraph{Acknowledgments}
+This work was supported by NSF award 2245300.
+
+\clearpage
+\pagestyle{empty}
+
+\begin{figure}
+  \centering
+  \caption{\textbf{Validation of the benchmarks.}
+    \textbf{a.} Each series indicates, for a given species, the number of accepted PSMs as a function of PSM-level FDR.
+    \textbf{b.} The bar plot indicates the proportion of spectra that were accepted at 1\% PSM-level FDR per species.
+    \textbf{c.} Each histogram shows, for one species, the distribution of the proportion of total ion current that is matched by b- or y-ions per accepted PSM, using a matching tolerance of 0.05~\textit{m/z}.}
+  \label{fig:fdr}
+\end{figure}
+
+\clearpage
+
+  \begin{tabular}{lll}
+    \includegraphics[height=4in]{nine-species-fdrs.pdf} &
+    \includegraphics[height=4in]{percent-id.pdf}  &
+    \includegraphics[height=4in]{match-by.pdf}  \\
+    a & b & c \\
+  \end{tabular}
+
 \end{document}

### 2024-10-01 doc/nine-species/nine-species.tex

@@
 \hline
-PXD005025 & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
-108,402 & 11,638 & 102,255 & 11,557
+PXD005025 \cite{paiva2016label} & \textit{Vigna mungo} & UP000087766 & 24 & 932,848 &
+\edit{108,402} & \edit{11,638} & \edit{102,255} & \edit{11,557}
 & 20 & 0.05 \\
-PXD004948 & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
-25,522 & 5630 & 25,522 & 5630
+PXD004948 \cite{nevo2017impact} & \textit{Mus musculus} & UP000000589 & 13 & 306,786 & 
+\edit{25,522} & \edit{5630} & \edit{25,522} & \edit{5630}
 & 10 & 0.05 \\
-PXD004325 & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
-267,183 & 15,220 & 100,485 & 11,934
+PXD004325 \cite{cassidy2016combination} & \textit{Methanosarcina mazei} & UP000033058  & 72 & 3,728,183 &
+\edit{267,183} & \edit{15,220} & \edit{100,485} & \edit{11,934}
 & 10 & 0.05 \\
-PXD004565 & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
-1,351,938 & 28,364 & 113,234 & 17,481
+PXD004565 \cite{reuss2017large} & \textit{Bacillus subtilis} & UP000001570 & 106 & 4,336,428 &
+\edit{1,351,938} & \edit{28,364} & \edit{113,234} & \edit{17,481}
 & 30 & 0.05 \\
-PXD004536 & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
-82,514 & 8080 & 82,514 & 8080
+PXD004536 \cite{petersen2016chemosynthetic} & \textit{Candidatus endoloripes} & UP000094849 & 11 & 2,272,023 &
+\edit{82,514} & \edit{8080} & \edit{82,514} & \edit{8080}
 & 20 & 0.05 \\
-PXD004947 & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
-177,553 & 48,459 & 100,056 & 35,787
+PXD004947 \cite{mata2017depth} & \textit{Solanum lycopersicum} & UP000004994 & 60 & 603,506 &
+\edit{177,553} & \edit{48,459} & \edit{100,056} & \edit{35,787}
 & 15 & 0.05 \\
-PXD003868 & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
-585,846 & 19,102 & 108,973 & 13,285
+PXD003868 \cite{seidel2017quantitative} & \textit{Saccharomyces-cerevisiae} & UP000002311 & 27 & 1,477,397 &
+\edit{585,846} & \edit{19,102} & \edit{108,973} & \edit{13,285}
 & 20 & 0.05 \\
-PXD004467 & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
-194,604 & 21,081 & 102,285 & 18,630
+PXD004467 \cite{hu2016proteome} & \textit{Apis mellifera} & UP000005203 & 17 & 823,169 &
+\edit{194,604} & \edit{21,081} & \edit{102,285} & \edit{18,630}
 & 20 & 0.05 \\
-PXD004424 & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
-44,555 & 10,848 & 44,555 & 10,848
+PXD004424 \cite{cypryk2017proteomic} & \textit{H.\ sapiens} & UP000005640 & 26 & 684,821 &
+\edit{44,555} & \edit{10,848} & \edit{44,555} & \edit{10,848}
 & 20 & 0.02 \\
